---
title: "Carbonate-derived Inorganic Carbon (DIC)"
author: "Xinlin Ni"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, warning = FALSE, message = FALSE,
  fig.width = 10, fig.height = 7, dpi = 300
)
```

# 1. Analysis overview

This workflow analyses carbonate-derived inorganic carbon across enhanced rock
weathering (ERW) application levels, water types, and sampling dates.

The primary analysis first compares centred raw ERW with centred
`log1p(ERW)` using identical additive models fitted by maximum likelihood
(ML). If the two forms differ by less than 2 AIC units, raw ERW is retained for
simpler interpretation; otherwise, the lower-AIC form is retained. Only after
the ERW form is selected is the `water_type * ERW` interaction compared with
the additive structure. The selected final model is refitted using restricted
maximum likelihood (REML). Random intercepts account for block, mesocosm nested
within block, and sampling date.

Figures retain the original ERW values as axis labels. Any square-root axis
transformation is for display only and does not determine the modelled ERW
functional form.

Date-specific analyses inherit the ERW form selected by the overall model, then
compare additive and interaction structures using ML before REML refitting.
Within a single date, each mesocosm contributes only one observation, so the
daily models include block—but not mesocosm—as a random intercept.

# 2. Load packages

```{r packages}
#rm(list = ls())

library(tidyverse)
library(readxl)
library(lme4)
library(lmerTest)
library(emmeans)
library(performance)
library(parameters)
library(broom.mixed)
library(janitor)
library(patchwork)
library(scales)
```

# 3. Analysis settings

```{r settings}
DATA_FILE <- "../data/ERW_IC_results_R_ready.xlsx"

if (is.na(DATA_FILE)) {
  stop(
    "Could not find ERW_IC_results_R_ready.xlsx. ",
    "Place it beside this Rmd, inside a data/ folder, or update DATA_FILE_CANDIDATES."
  )
}

DATA_SHEET <- "ic_data"
ERW_LEVELS <- c(0, 5, 10, 20, 50, 100)
ERW_RAW_CENTRE <- mean(ERW_LEVELS)
ERW_LOG_CENTRE <- mean(log1p(ERW_LEVELS))
AIC_EQUIVALENCE_THRESHOLD <- 2

WATER_COLOURS <- c("Ground" = "#0072B2", "River" = "#D55E00")
WATER_LABELS <- c("Ground" = "Groundwater", "River" = "River water")
WATER_SHAPES <- c("Ground" = 16, "River" = 17)

OUTPUT_DIR <- "../outputs/DIC"
dir.create(OUTPUT_DIR, showWarnings = FALSE)
```

# 4. Import the data

```{r import-data}
message("Reading data from: ", normalizePath(DATA_FILE))

ic_raw <- read_excel(
  path = DATA_FILE,
  sheet = DATA_SHEET
)

glimpse(ic_raw)
```

# 5. Prepare the carbonate dataset

## 5.1 Select and format variables

```{r select-format-variables}
carbonate_dat <- ic_raw %>%
  transmute(
    sample_id,
    sample_key,
    sample_date = as.Date(sample_date),
    mesocosm = factor(mesocosm),
    block = factor(block),
    water_type = factor(water_type, levels = c("Ground", "River")),
    ERW_level = as.numeric(erw_level),
    carbonate_mg_l = as.numeric(carbonate_mg_l),
    qa_flag
  )
```

## 5.2 Convert carbonate to inorganic carbon

Ion chromatography results are converted from mg carbonate per litre to mg
carbon per litre using:

\[
\mathrm{mg\ C\ L^{-1}}
=
\mathrm{mg\ CO_3^{2-}\ L^{-1}}
\times
\frac{12.011}{60.008}
\]

```{r convert-carbonate}
carbonate_dat <- carbonate_dat %>%
  mutate(
    inorganic_C_mg_l = carbonate_mg_l * (12.011 / 60.008)
  )
```

The conversion changes the response units and coefficient scale, but not the
fitted pattern or p-values relative to the equivalent carbonate model.

## 5.3 Create model variables

```{r create-model-variables}
carbonate_dat <- carbonate_dat %>%
  mutate(
    sample_date_f = factor(sample_date, levels = sort(unique(sample_date))),
    days_since_start = as.numeric(sample_date - min(sample_date, na.rm = TRUE)),
    erw_raw_c = ERW_level - ERW_RAW_CENTRE,
    erw_log = log1p(ERW_level),
    erw_log_c = erw_log - ERW_LOG_CENTRE
  )
```

`erw_raw_c` is the centred but otherwise untransformed ERW application rate.
`erw_log_c` is centred `log(1 + ERW_level)`. Centring changes the interpretation
of the intercept but does not change fitted values or the functional shape.

```{r inspect-derived-variables}
ERW_LOG_CENTRE

carbonate_dat %>%
  distinct(ERW_level, erw_raw_c, erw_log, erw_log_c) %>%
  arrange(ERW_level)

carbonate_dat %>%
  distinct(sample_date, sample_date_f, days_since_start) %>%
  arrange(sample_date)
```

# 6. Data checks

```{r data-summary}
carbonate_dat %>%
  summarise(
    total_rows = n(),
    number_of_mesocosms = n_distinct(mesocosm),
    number_of_blocks = n_distinct(block),
    number_of_dates = n_distinct(sample_date),
    missing_carbonate = sum(is.na(carbonate_mg_l)),
    missing_inorganic_C = sum(is.na(inorganic_C_mg_l)),
    flagged_rows = sum(!is.na(qa_flag))
  )
```

```{r treatment-structure}
carbonate_dat %>%
  count(sample_date, water_type, ERW_level) %>%
  arrange(sample_date, water_type, ERW_level)
```

```{r duplicate-check}
carbonate_dat %>%
  count(sample_date, mesocosm) %>%
  filter(n > 1)
```

```{r qa-flags}
carbonate_dat %>%
  filter(!is.na(qa_flag)) %>%
  select(
    sample_id, sample_date, mesocosm, block, water_type,
    ERW_level, carbonate_mg_l, inorganic_C_mg_l, qa_flag
  )
```

QA-flagged rows are not removed automatically. Exclude a row only when its QA
information indicates that the carbonate measurement itself is unreliable.

# 7. Create the model dataset

```{r create-model-data}
carbonate_model_dat <- carbonate_dat %>%
  filter(
    !is.na(inorganic_C_mg_l),
    !is.na(water_type),
    !is.na(ERW_level),
    !is.na(erw_raw_c),
    !is.na(erw_log_c),
    !is.na(sample_date_f),
    !is.na(block),
    !is.na(mesocosm)
  ) %>%
  droplevels()
```

```{r inspect-model-data}
carbonate_model_dat %>%
  summarise(
    n_observations = n(),
    n_mesocosms = n_distinct(mesocosm),
    n_blocks = n_distinct(block),
    n_dates = n_distinct(sample_date)
  )
```

# 8. Descriptive statistics

```{r descriptive-summary}
carbonate_summary <- carbonate_model_dat %>%
  group_by(sample_date, sample_date_f, water_type, ERW_level) %>%
  summarise(
    n = n(),
    mean_inorganic_C = mean(inorganic_C_mg_l, na.rm = TRUE),
    sd_inorganic_C = sd(inorganic_C_mg_l, na.rm = TRUE),
    se_inorganic_C = sd_inorganic_C / sqrt(n),
    .groups = "drop"
  )

carbonate_summary
```

```{r descriptive-summary}
# First sampling date after ERW application, and last sampling date
first_post_erw_date <- carbonate_dat %>%
  filter(sample_date >= as.Date("2025-05-15")) %>%
  summarise(min(sample_date)) %>%
  pull()

last_sample_date <- max(carbonate_dat$sample_date, na.rm = TRUE)

first_post_erw_date
last_sample_date

# Mean DIC at first post-application sampling and at last sampling,
# by ERW level (groundwater and river water combined)
dic_decline_table <- carbonate_dat %>%
  filter(sample_date %in% c(first_post_erw_date, last_sample_date)) %>%
  mutate(
    timepoint = if_else(
      sample_date == first_post_erw_date,
      "first_post_erw",
      "last_sample"
    )
  ) %>%
  group_by(ERW_level, timepoint) %>%
  summarise(
    mean_DIC = mean(inorganic_C_mg_l, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_wider(
    names_from = timepoint,
    values_from = mean_DIC
  ) %>%
  mutate(
    decline = first_post_erw - last_sample,
    decline_pct = decline / first_post_erw * 100
  ) %>%
  arrange(ERW_level)

dic_decline_table
```

# 9. Exploratory figures

All ERW figures display the original treatment values, but position them using
a square-root x-axis transformation. This retains the zero treatment and makes the
treatment spacing easier to inspect.

## 9.1 Raw observations with means ± SD

```{r combined-raw-mean-sd}
water_offset <- 0.12

carbonate_summary_plot <- carbonate_summary %>%
  mutate(
    x_plot = sqrt(ERW_level) + if_else(
      water_type == "Ground",
      -water_offset,
      water_offset
    )
  )

carbonate_model_plot <- carbonate_model_dat %>%
  mutate(
    x_plot = sqrt(ERW_level) + if_else(
      water_type == "Ground",
      -water_offset,
      water_offset
    )
  )

p_combined <- ggplot(
  carbonate_model_plot,
  aes(
    x = x_plot,
    y = inorganic_C_mg_l,
    colour = water_type,
    shape = water_type
  )
) +
  geom_point(
    position = position_jitter(width = 0.035, height = 0),
    alpha = 0.35,
    size = 1.5
  ) +
  geom_line(
    data = carbonate_summary_plot,
    aes(y = mean_inorganic_C, group = water_type),
    linewidth = 0.8
  ) +
  geom_point(
    data = carbonate_summary_plot,
    aes(y = mean_inorganic_C),
    size = 2.5,
    stroke = 0.8
  ) +
  geom_errorbar(
    data = carbonate_summary_plot,
    aes(
      y = mean_inorganic_C,
      ymin = pmax(mean_inorganic_C - sd_inorganic_C, 0),
      ymax = mean_inorganic_C + sd_inorganic_C
    ),
    width = 0.07,
    linewidth = 0.6
  ) +
  facet_wrap(~ sample_date_f, ncol = 4) +
  scale_x_continuous(
    breaks = sqrt(ERW_LEVELS),
    labels = ERW_LEVELS
  ) +
  scale_colour_manual(values = WATER_COLOURS, labels = WATER_LABELS) +
  scale_shape_manual(values = WATER_SHAPES, labels = WATER_LABELS) +
  labs(
    x = expression("ERW application (t ha"^{-1}*")"),
    y = expression("DIC  (mg C L"^{-1}*")"),
    colour = "Water type",
    shape = "Water type"
  ) +
  theme_classic() +
  theme(
    legend.position = "right",
    legend.direction = "vertical",
    strip.background = element_rect(fill = "grey95", colour = "grey70"),
    strip.text = element_text(face = "bold", size = 13),
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 13),
    legend.title = element_text(size = 13),
    legend.text = element_text(size = 13)
  )

p_combined
```

## 9.2 Zero-ERW control time series

```{r control-summary}
control_summary <- carbonate_model_dat %>%
  filter(ERW_level == 0) %>%
  group_by(sample_date, water_type) %>%
  summarise(
    n = n(),
    mean_inorganic_C = mean(inorganic_C_mg_l, na.rm = TRUE),
    sd_inorganic_C = sd(inorganic_C_mg_l, na.rm = TRUE),
    .groups = "drop"
  )
```

```{r control-timeseries}
p_control <- ggplot(
  control_summary,
  aes(
    x = sample_date,
    y = mean_inorganic_C,
    colour = water_type,
    shape = water_type,
    group = water_type
  )
) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2.8) +
  geom_errorbar(
    aes(
      ymin = pmax(mean_inorganic_C - sd_inorganic_C, 0),
      ymax = mean_inorganic_C + sd_inorganic_C
    ),
    width = 0.5,
    linewidth = 0.6
  ) +
  scale_colour_manual(values = WATER_COLOURS, labels = WATER_LABELS) +
  scale_shape_manual(values = WATER_SHAPES, labels = WATER_LABELS) +
  labs(
    x = "Sampling date",
    y = expression("0 t/ha inorganic carbon (mg C L"^{-1}*")"),
    colour = "Water type",
    shape = "Water type"
  ) +
  theme_classic() +
  theme(legend.position = "top")

p_control
```

# 10. Select the ERW form, then the fixed-effect structure

All fixed-effect comparisons use ML rather than REML. First, raw and log1p ERW
are compared with the same additive structure. The interaction is tested only
after selecting the ERW form.

Samples collected on 14 May 2025 are retained in the descriptive summaries and
mean ± SD figures above, but excluded from all models because they were
collected before ERW products were applied.

```{r exclude-pre-erw-data-from-models}
carbonate_model_dat_post_erw <- carbonate_model_dat %>%
  filter(sample_date != as.Date("2025-05-14")) %>%
  droplevels()

carbonate_model_dat_post_erw %>%
  filter(ERW_level == 0) %>%
  summarise(
    mean_control_DIC = mean(inorganic_C_mg_l, na.rm = TRUE)
  )
```

## 10.1 Compare raw and log1p ERW using additive ML models

```{r fit-primary-candidate-models}
erw_raw_additive_ml <- lmer(
  inorganic_C_mg_l ~
    water_type +
    erw_raw_c +
    (1 | block/mesocosm) +
    (1 | sample_date_f),
  data = carbonate_model_dat_post_erw,
  REML = FALSE,
  na.action = na.exclude,
  control = lmerControl(
    optimizer = "bobyqa",
    optCtrl = list(maxfun = 200000)
  )
)

erw_log1p_additive_ml <- lmer(
  inorganic_C_mg_l ~
    water_type +
    erw_log_c +
    (1 | block/mesocosm) +
    (1 | sample_date_f),
  data = carbonate_model_dat_post_erw,
  REML = FALSE,
  na.action = na.exclude,
  control = lmerControl(
    optimizer = "bobyqa",
    optCtrl = list(maxfun = 200000)
  )
)

erw_form_comparison <- tibble(
  erw_form = c("raw", "log1p"),
  fixed_structure = "water_type + ERW",
  n = c(
    nobs(erw_raw_additive_ml),
    nobs(erw_log1p_additive_ml)
  ),
  logLik = c(
    as.numeric(logLik(erw_raw_additive_ml)),
    as.numeric(logLik(erw_log1p_additive_ml))
  ),
  AIC = c(
    AIC(erw_raw_additive_ml),
    AIC(erw_log1p_additive_ml)
  ),
  BIC = c(
    BIC(erw_raw_additive_ml),
    BIC(erw_log1p_additive_ml)
  )
) %>%
  arrange(AIC) %>%
  mutate(delta_AIC = AIC - min(AIC))

aic_difference <- diff(range(erw_form_comparison$AIC))

selected_erw_form <- if (
  aic_difference < AIC_EQUIVALENCE_THRESHOLD
) {
  "raw"
} else {
  erw_form_comparison$erw_form[
    which.min(erw_form_comparison$AIC)
  ]
}

erw_form_comparison <- erw_form_comparison %>%
  mutate(
    selected_final = erw_form == selected_erw_form
  )

erw_form_candidate_fixed_effects <- bind_rows(
  broom.mixed::tidy(
    erw_raw_additive_ml,
    effects = "fixed",
    conf.int = TRUE
  ) %>% mutate(erw_form = "raw", .before = 1),
  broom.mixed::tidy(
    erw_log1p_additive_ml,
    effects = "fixed",
    conf.int = TRUE
  ) %>% mutate(erw_form = "log1p", .before = 1)
)

erw_form_candidate_type2_tests <- bind_rows(
  anova(
    erw_raw_additive_ml,
    type = 2,
    ddf = "Satterthwaite"
  ) %>%
    as.data.frame() %>%
    rownames_to_column("term") %>%
    clean_names() %>%
    mutate(erw_form = "raw", .before = 1),
  anova(
    erw_log1p_additive_ml,
    type = 2,
    ddf = "Satterthwaite"
  ) %>%
    as.data.frame() %>%
    rownames_to_column("term") %>%
    clean_names() %>%
    mutate(erw_form = "log1p", .before = 1)
)

erw_form_comparison
erw_form_candidate_fixed_effects
erw_form_candidate_type2_tests
selected_erw_form
```

## 10.2 Create the selected ERW predictor and test the interaction

```{r compare-primary-models}
carbonate_model_dat_post_erw <- carbonate_model_dat_post_erw %>%
  mutate(
    erw_x = if (selected_erw_form == "raw") {
      erw_raw_c
    } else {
      erw_log_c
    }
  )

if (!any(carbonate_model_dat_post_erw$ERW_level == 0, na.rm = TRUE)) {
  stop("The post-ERW model dataset does not contain the 0 t ha^-1 control.")
}

selected_model_treatment_balance <- carbonate_model_dat_post_erw %>%
  count(ERW_level, name = "n")

selected_model_treatment_balance

primary_additive_ml <- lmer(
  inorganic_C_mg_l ~
    water_type +
    erw_x +
    (1 | block/mesocosm) +
    (1 | sample_date_f),
  data = carbonate_model_dat_post_erw,
  REML = FALSE,
  na.action = na.exclude,
  control = lmerControl(
    optimizer = "bobyqa",
    optCtrl = list(maxfun = 200000)
  )
)

primary_interaction_ml <- lmer(
  inorganic_C_mg_l ~
    water_type * erw_x +
    (1 | block/mesocosm) +
    (1 | sample_date_f),
  data = carbonate_model_dat_post_erw,
  REML = FALSE,
  na.action = na.exclude,
  control = lmerControl(
    optimizer = "bobyqa",
    optCtrl = list(maxfun = 200000)
  )
)

primary_model_comparison <- anova(
  primary_additive_ml,
  primary_interaction_ml
)

primary_model_comparison

primary_interaction_p <- primary_model_comparison$`Pr(>Chisq)`[2]
primary_interaction_p
```

## 10.3 Refit the selected structure using REML

```{r refit-primary-model}
primary_model_structure <- if (
  !is.na(primary_interaction_p) && primary_interaction_p < 0.05
) {
  "interaction"
} else {
  "additive"
}

model_dic_final <- if (primary_model_structure == "interaction") {
  update(primary_interaction_ml, REML = TRUE)
} else {
  update(primary_additive_ml, REML = TRUE)
}

primary_model_structure
summary(model_dic_final)
```

# 11. Primary model results

## 11.1 Type II fixed-effect tests

```{r primary-model-anova}
model_dic_anova <- anova(
  model_dic_final,
  type = 2,
  ddf = "Satterthwaite"
)

model_dic_anova
```

## 11.2 Fixed-effect coefficients

```{r primary-fixed-effects}
fixed_effects <- broom.mixed::tidy(
  model_dic_final,
  effects = "fixed",
  conf.int = TRUE
)

fixed_effects
```

## 11.3 Random-effect estimates

```{r primary-random-effects}
random_effects <- broom.mixed::tidy(
  model_dic_final,
  effects = "ran_pars",
  conf.int = TRUE
)

random_effects
```

## 11.4 Model performance

```{r primary-model-performance}
model_r2 <- performance::r2_nakagawa(model_dic_final)
model_icc <- performance::icc(model_dic_final)
model_information <- performance::model_performance(model_dic_final)

model_r2
model_icc
model_information
```

## 11.5 Technical checks

```{r primary-technical-checks}
lme4::isSingular(model_dic_final, tol = 1e-4)
performance::check_collinearity(model_dic_final)
```

# 12. Estimated marginal means and overall predictions

## 12.1 Water-type estimated marginal means

```{r water-type-emmeans}
water_type_emmeans <- emmeans(
  model_dic_final,
  ~ water_type
) %>%
  as.data.frame()

water_type_emmeans
```

## 12.2 Overall ERW slope

For an additive model, one common ERW slope is estimated across both water
types. If the interaction is retained, separate slopes are estimated.

```{r overall-erw-slopes}
overall_erw_slopes <- if (primary_model_structure == "interaction") {
  emtrends(
    model_dic_final,
    ~ water_type,
    var = "erw_x"
  ) %>%
    summary(infer = c(TRUE, TRUE)) %>%
    as.data.frame()
} else {
  emtrends(
    model_dic_final,
    ~ 1,
    var = "erw_x"
  ) %>%
    summary(infer = c(TRUE, TRUE)) %>%
    as.data.frame()
}

overall_erw_slopes
```

If `selected_erw_form` is `"raw"`, the ERW coefficient represents the change in
inorganic carbon for a 1 t ha^-1 increase in ERW application. If it is `"log1p"`,
the coefficient represents the change associated with a one-unit increase in
`log(1 + ERW)`; effects per raw ERW unit are therefore not constant in that model.

## 12.3 Population-level predictions

```{r overall-prediction-data}
emm_dic <- emmeans(
  model_dic_final,
  ~ water_type | erw_x,
  at = list(
    erw_x = if (selected_erw_form == "raw") {
      ERW_LEVELS - ERW_RAW_CENTRE
    } else {
      log1p(ERW_LEVELS) - ERW_LOG_CENTRE
    }
  )
)

emm_dic_dat <- emm_dic %>%
  as.data.frame() %>%
  mutate(
    ERW_level = if (selected_erw_form == "raw") {
      erw_x + ERW_RAW_CENTRE
    } else {
      expm1(erw_x + ERW_LOG_CENTRE)
    }
  )

emm_dic_dat

```

```{r overall-prediction-figure}
p_predictions <- ggplot(
  emm_dic_dat,
  aes(
    x = ERW_level,
    y = emmean,
    colour = water_type,
    shape = water_type,
    group = water_type
  )
) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2.7) +
  geom_errorbar(
    aes(ymin = lower.CL, ymax = upper.CL),
    width = 0.07,
    linewidth = 0.6
  ) +
  scale_x_continuous(
    trans = scales::transform_sqrt(),
    breaks = ERW_LEVELS
  ) +
  scale_colour_manual(values = WATER_COLOURS, labels = WATER_LABELS) +
  scale_shape_manual(values = WATER_SHAPES, labels = WATER_LABELS) +
  labs(
    x = expression("ERW application (t ha"^{-1}*")"),
    y = expression("Model-estimated inorganic carbon (mg C L"^{-1}*")"),
    colour = "Water type",
    shape = "Water type"
  ) +
  theme_classic() +
  theme(legend.position = "top")

p_predictions
```

# 13. Date-specific supplementary models

The overall mixed model remains the primary analysis. Within a single date,
mesocosm and sampling date cannot be fitted as random effects because each
mesocosm supplies only one observation and date is constant. Block is retained
as the only random intercept.

For every date:

1. use the ERW form selected by the overall analysis;
2. fit interaction and additive models using ML;
3. compare them using a likelihood-ratio test;
4. retain the interaction only when `p < 0.05`;
5. refit the selected structure using REML.

## 13.1 Daily model-selection function

```{r daily-model-function}
fit_daily_dic_models <- function(dat) {

  interaction_ml <- lmer(
    inorganic_C_mg_l ~
      water_type * erw_x +
      (1 | block),
    data = dat,
    REML = FALSE,
    na.action = na.exclude,
    control = lmerControl(
      optimizer = "bobyqa",
      optCtrl = list(maxfun = 200000)
    )
  )

  additive_ml <- lmer(
    inorganic_C_mg_l ~
      water_type +
      erw_x +
      (1 | block),
    data = dat,
    REML = FALSE,
    na.action = na.exclude,
    control = lmerControl(
      optimizer = "bobyqa",
      optCtrl = list(maxfun = 200000)
    )
  )

  comparison <- anova(additive_ml, interaction_ml)
  interaction_p <- comparison$`Pr(>Chisq)`[2]

  selected_structure <- if (
    !is.na(interaction_p) && interaction_p < 0.05
  ) {
    "interaction"
  } else {
    "additive"
  }

  final_model <- if (selected_structure == "interaction") {
    update(interaction_ml, REML = TRUE)
  } else {
    update(additive_ml, REML = TRUE)
  }

  list(
    additive_ml = additive_ml,
    interaction_ml = interaction_ml,
    comparison = comparison,
    interaction_p = interaction_p,
    selected_structure = selected_structure,
    final_model = final_model
  )
}
```

## 13.2 Fit models for every date

```{r fit-daily-models}
daily_models <- carbonate_model_dat_post_erw %>%
  group_by(sample_date, sample_date_f) %>%
  nest() %>%
  mutate(
    fits = map(data, fit_daily_dic_models),
    interaction_p = map_dbl(fits, "interaction_p"),
    selected_structure = map_chr(fits, "selected_structure"),
    model = map(fits, "final_model")
  )

daily_models %>%
  select(sample_date, sample_date_f, interaction_p, selected_structure)
```

## 13.3 Daily model comparisons

```{r daily-model-comparisons}
daily_model_comparison <- daily_models %>%
  transmute(
    sample_date,
    sample_date_f,
    comparison = map(fits, "comparison")
  ) %>%
  mutate(
    comparison = map(
      comparison,
      ~ as.data.frame(.x) %>%
        rownames_to_column("candidate_model") %>%
        clean_names()
    )
  ) %>%
  unnest(comparison)

daily_model_comparison
```

```{r daily-selection-summary}
daily_selection_summary <- daily_models %>%
  transmute(
    sample_date,
    sample_date_f,
    interaction_p,
    selected_structure,
    interaction_supported = interaction_p < 0.05
  )

daily_selection_summary
```

## 13.4 Daily fixed-effect coefficients

```{r daily-fixed-effects}
daily_coefficients <- daily_models %>%
  mutate(
    coefficients = map(
      model,
      ~ broom.mixed::tidy(
        .x,
        effects = "fixed",
        conf.int = TRUE
      )
    )
  ) %>%
  select(
    sample_date,
    sample_date_f,
    selected_structure,
    coefficients
  ) %>%
  unnest(coefficients) %>%
  group_by(term) %>%
  mutate(
    p_value_fdr = if_else(
      term == "(Intercept)",
      NA_real_,
      p.adjust(p.value, method = "BH")
    )
  ) %>%
  ungroup()

daily_coefficients
```

## 13.5 Daily Type II tests

```{r daily-term-tests}
daily_term_tests <- daily_models %>%
  mutate(
    term_tests = map(
      model,
      ~ anova(
        .x,
        type = 2,
        ddf = "Satterthwaite"
      ) %>%
        as.data.frame() %>%
        rownames_to_column("term") %>%
        clean_names()
    )
  ) %>%
  select(
    sample_date,
    sample_date_f,
    selected_structure,
    term_tests
  ) %>%
  unnest(term_tests) %>%
  group_by(term) %>%
  mutate(
    p_value_fdr = p.adjust(pr_f, method = "BH")
  ) %>%
  ungroup()

daily_term_tests
```

FDR adjustment is applied across dates separately for each tested term.

## 13.6 Concise daily ERW coefficient table

For an additive model, `erw_x` is the common ERW slope. If an interaction
is selected, the reference-water slope and interaction coefficient must be
interpreted together.

```{r daily-erw-coefficients}
daily_erw_coefficients <- daily_coefficients %>%
  filter(str_detect(term, "erw_x")) %>%
  mutate(
    significance = case_when(
      p_value_fdr < 0.001 ~ "***",
      p_value_fdr < 0.01 ~ "**",
      p_value_fdr < 0.05 ~ "*",
      p_value_fdr < 0.10 ~ ".",
      TRUE ~ "ns"
    )
  )

daily_erw_coefficients
```

## 13.7 Daily model predictions

```{r daily-prediction-data}
daily_predictions <- daily_models %>%
  mutate(
    predictions = map(
      model,
      ~ emmeans(
        .x,
        ~ water_type | erw_x,
        at = list(
          erw_x = if (selected_erw_form == "raw") {
            ERW_LEVELS - ERW_RAW_CENTRE
          } else {
            log1p(ERW_LEVELS) - ERW_LOG_CENTRE
          }
        )
      ) %>%
        as.data.frame() %>%
        mutate(
          ERW_level = if (selected_erw_form == "raw") {
            erw_x + ERW_RAW_CENTRE
          } else {
            expm1(erw_x + ERW_LOG_CENTRE)
          }
        )
    )
  ) %>%
  select(
    sample_date,
    sample_date_f,
    selected_structure,
    predictions
  ) %>%
  unnest(predictions)

daily_predictions
```

```{r daily-prediction-figure}
p_daily_predictions <- ggplot(
  daily_predictions,
  aes(
    x = ERW_level,
    y = emmean,
    colour = water_type,
    shape = water_type,
    group = water_type
  )
) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2.4) +
  geom_errorbar(
    aes(ymin = lower.CL, ymax = upper.CL),
    width = 0.07,
    linewidth = 0.5
  ) +
  facet_wrap(~ sample_date_f, ncol = 4) +
  scale_x_continuous(
    trans = scales::transform_sqrt(),
    breaks = ERW_LEVELS
  ) +
  scale_colour_manual(values = WATER_COLOURS, labels = WATER_LABELS) +
  scale_shape_manual(values = WATER_SHAPES, labels = WATER_LABELS) +
  labs(
    x = expression("ERW application (t ha"^{-1}*")"),
    y = expression("Date-specific model estimate (mg C L"^{-1}*")"),
    colour = "Water type",
    shape = "Water type"
  ) +
  theme_classic() +
  theme(
    legend.position = "top",
    strip.background = element_rect(fill = "grey95", colour = "grey70"),
    strip.text = element_text(face = "bold")
  )

p_daily_predictions
```

## 13.8 Daily model technical checks

```{r daily-model-checks}
daily_model_checks <- daily_models %>%
  transmute(
    sample_date,
    sample_date_f,
    selected_structure,
    singular_fit = map_lgl(
      model,
      ~ lme4::isSingular(.x, tol = 1e-4)
    ),
    block_variance = map_dbl(
      model,
      ~ {
        vc <- as.data.frame(VarCorr(.x))
        value <- vc %>%
          filter(grp == "block", var1 == "(Intercept)") %>%
          pull(vcov)
        if (length(value) == 0) NA_real_ else value[1]
      }
    )
  )

daily_model_checks
```

A singular daily model means that block variance is estimated at or near zero
for that date. It does not justify adding mesocosm as a random effect.

# 14. Primary model diagnostics

## 14.1 Automated diagnostic panel

```{r automated-diagnostics}
performance::check_model(model_dic_final)
```

## 14.2 Residual diagnostic plots

```{r residual-diagnostics}
old_par <- par(no.readonly = TRUE)
par(mfrow = c(2, 2))

plot(
  fitted(model_dic_final),
  resid(model_dic_final),
  xlab = "Fitted values",
  ylab = "Residuals",
  main = "Residuals versus fitted values"
)
abline(h = 0, lty = 2)

qqnorm(resid(model_dic_final), main = "Normal Q-Q plot")
qqline(resid(model_dic_final))

hist(
  resid(model_dic_final),
  main = "Residual distribution",
  xlab = "Residual"
)

plot(
  carbonate_model_dat_post_erw$inorganic_C_mg_l,
  fitted(model_dic_final),
  xlab = "Observed values",
  ylab = "Fitted values",
  main = "Observed versus fitted"
)
abline(0, 1, lty = 2)

par(old_par)
```

# 15. Model summary table

```{r primary-model-fit-table}
model_fit <- tibble(
  response = "inorganic_C_mg_l",
  selected_fixed_structure = primary_model_structure,
  fixed_effects = if_else(
    primary_model_structure == "interaction",
    "water_type * erw_x",
    "water_type + erw_x"
  ),
  random_effects =
    "block + mesocosm nested within block + sample_date",
  ERW_transformation = selected_erw_form,
  ERW_centre = if_else(
    selected_erw_form == "raw",
    ERW_RAW_CENTRE,
    ERW_LOG_CENTRE
  ),
  AIC = AIC(model_dic_final),
  BIC = BIC(model_dic_final),
  log_likelihood = as.numeric(logLik(model_dic_final)),
  n_observations = nobs(model_dic_final),
  n_mesocosms = n_distinct(carbonate_model_dat_post_erw$mesocosm),
  n_blocks = n_distinct(carbonate_model_dat_post_erw$block),
  n_dates = n_distinct(carbonate_model_dat_post_erw$sample_date),
  singular_fit = lme4::isSingular(model_dic_final, tol = 1e-4)
)

model_fit
```

# 16. Export tables and figures

## 16.1 Export tables

```{r export-tables}
write_csv(carbonate_summary, file.path(OUTPUT_DIR, "IC_summary_mean_sd.csv"))

write_csv(erw_form_comparison,
          file.path(OUTPUT_DIR, "IC_ERW_form_comparison.csv"))
write_csv(erw_form_candidate_fixed_effects,
          file.path(OUTPUT_DIR, "IC_ERW_form_candidate_fixed_effects.csv"))
write_csv(erw_form_candidate_type2_tests,
          file.path(OUTPUT_DIR, "IC_ERW_form_candidate_type2_tests.csv"))
write_csv(selected_model_treatment_balance,
          file.path(OUTPUT_DIR, "IC_selected_model_ERW_level_counts.csv"))

write_csv(
  as.data.frame(model_dic_anova) %>% rownames_to_column("term"),
  file.path(OUTPUT_DIR, "IC_final_model_type2_tests.csv")
)

write_csv(fixed_effects, file.path(OUTPUT_DIR, "IC_final_model_fixed_effects.csv"))
write_csv(random_effects, file.path(OUTPUT_DIR, "IC_final_model_random_effects.csv"))

write_csv(
  as.data.frame(primary_model_comparison) %>%
    rownames_to_column("candidate_model"),
  file.path(OUTPUT_DIR, "IC_selected_ERW_interaction_comparison.csv")
)

write_csv(model_fit, file.path(OUTPUT_DIR, "IC_final_model_fit.csv"))
write_csv(water_type_emmeans, file.path(OUTPUT_DIR, "IC_water_type_emmeans.csv"))
write_csv(overall_erw_slopes, file.path(OUTPUT_DIR, "IC_overall_ERW_slopes.csv"))
write_csv(emm_dic_dat, file.path(OUTPUT_DIR, "IC_final_model_predictions.csv"))
write_csv(daily_model_comparison, file.path(OUTPUT_DIR, "IC_daily_model_comparison.csv"))
write_csv(daily_selection_summary, file.path(OUTPUT_DIR, "IC_daily_model_selection.csv"))
write_csv(daily_coefficients, file.path(OUTPUT_DIR, "IC_daily_model_coefficients.csv"))
write_csv(daily_term_tests, file.path(OUTPUT_DIR, "IC_daily_model_term_tests.csv"))
write_csv(daily_erw_coefficients, file.path(OUTPUT_DIR, "IC_daily_ERW_coefficients.csv"))
write_csv(daily_predictions, file.path(OUTPUT_DIR, "IC_daily_model_predictions.csv"))
write_csv(daily_model_checks, file.path(OUTPUT_DIR, "IC_daily_model_checks.csv"))
```

## 16.2 Export figures

```{r export-figures}
ggsave(
  file.path(OUTPUT_DIR, "IC_combined_raw_mean_SD_sqrt_x.png"),
  p_combined, width = 15, height = 7, dpi = 300
)

ggsave(
  file.path(OUTPUT_DIR, "IC_control_time_series.png"),
  p_control, width = 9, height = 5.5, dpi = 300
)

ggsave(
  file.path(OUTPUT_DIR, "IC_final_model_predictions_sqrt_x.png"),
  p_predictions, width = 9, height = 5.5, dpi = 300
)

ggsave(
  file.path(OUTPUT_DIR, "IC_daily_model_predictions_sqrt_x.png"),
  p_daily_predictions, width = 12, height = 7, dpi = 300
)
```

# 17. Session information

```{r session-information}
sessionInfo()
```
