---
title: "Enhanced Rock Weathering Flume Experiment: Dissolved Organic Carbon"
author: "Xinlin Ni"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    theme: flatly
params:
  doc_data_file: "../data/ERW_DOC.xlsx"
  doc_data_sheet: 1
  output_dir: "../outputs/DOC"
---

# Overview

DOC was sampled once, on 22 May 2025. All models include the complete ERW gradient (0, 5, 10, 20, 50 and 100 t ha\(^{-1}\)). Model selection proceeds in two stages: (1) compare centred raw ERW with centred `log1p(ERW)` in identical additive ML models; (2) test the water-type-by-ERW interaction using the selected ERW form. The final model is refitted by REML. Block is retained as a random intercept; date and mesocosm random effects are not fitted because there is one DOC observation per mesocosm.

# 1. Setup

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(readxl)
library(janitor)
library(lme4)
library(lmerTest)
library(performance)
library(emmeans)
library(broom.mixed)
library(writexl)
library(knitr)

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.align = "center", dpi = 300)
options(contrasts = c("contr.treatment", "contr.poly"))

ERW_LEVELS <- c(0, 5, 10, 20, 50, 100)
WATER_LEVELS <- c("Ground", "River")
WATER_LABELS <- c("Ground" = "Groundwater", "River" = "River water")
WATER_COLOURS <- c("Ground" = "#0072B2", "River" = "#D55E00")
WATER_SHAPES <- c("Ground" = 16, "River" = 17)
AIC_TIE_THRESHOLD <- 2
INTERACTION_ALPHA <- 0.05

DOC_THEME <- theme_classic(base_size = 11) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),
        plot.subtitle = element_text(hjust = 0.5, colour = "grey30"),
        axis.title = element_text(face = "bold"),
        legend.position = "right",
        legend.title = element_text(face = "bold"))

dir.create(params$output_dir, recursive = TRUE, showWarnings = FALSE)
```

# 2. Import and prepare data

```{r import-prepare}
if (!file.exists(params$doc_data_file)) {
  stop("DOC data file not found: ", params$doc_data_file,
       "\nEdit params$doc_data_file in the YAML header.")
}

doc_raw <- read_excel(params$doc_data_file, sheet = params$doc_data_sheet,
                      na = c("", "NA", "N/A", "n.a.")) %>%
  clean_names()

required_columns <- c("mesocosm", "block", "water_type", "erw_level", "date", "doc")
missing_columns <- setdiff(required_columns, names(doc_raw))
if (length(missing_columns) > 0) {
  stop("Missing required columns: ", paste(missing_columns, collapse = ", "))
}

doc_data <- doc_raw %>%
  transmute(
    Mesocosm = factor(mesocosm),
    Block = factor(block),
    Water_type = factor(water_type, levels = WATER_LEVELS),
    ERW_level = as.numeric(erw_level),
    ERW_log = log1p(ERW_level),
    Date = as.Date(date),
    DOC_mg_C_L = as.numeric(doc)
  ) %>%
  arrange(Water_type, ERW_level, Block)

glimpse(doc_data)
```

# 3. Checks and descriptive results

```{r checks-summary}
doc_checks <- tibble(
  check = c("Rows", "Mesocosms", "Blocks", "Sampling dates",
            "Missing DOC", "Duplicated mesocosms"),
  value = c(nrow(doc_data), n_distinct(doc_data$Mesocosm),
            n_distinct(doc_data$Block), n_distinct(doc_data$Date),
            sum(is.na(doc_data$DOC_mg_C_L)),
            sum(duplicated(doc_data$Mesocosm)))
)

doc_summary <- doc_data %>%
  group_by(Water_type, ERW_level) %>%
  summarise(n = sum(!is.na(DOC_mg_C_L)),
            mean = mean(DOC_mg_C_L, na.rm = TRUE),
            sd = sd(DOC_mg_C_L, na.rm = TRUE),
            minimum = min(DOC_mg_C_L, na.rm = TRUE),
            maximum = max(DOC_mg_C_L, na.rm = TRUE),
            .groups = "drop")

doc_checks %>% kable(caption = "DOC dataset checks")
doc_summary %>%
  mutate(Water_type = dplyr::recode(as.character(Water_type), !!!WATER_LABELS)) %>%
  kable(digits = 3, caption = "DOC descriptive statistics")

control_DOC_mean <- doc_data %>%
  filter(ERW_level == 0) %>%
  summarise(
    mean_control_DOC = mean(DOC_mg_C_L, na.rm = TRUE)
  )

control_DOC_mean
```

The square-root spacing below is only a visual display choice; it is unrelated to the ERW form used in the models. A separate 0 t ha\(^{-1}\) plot is unnecessary because the controls are included here and in all models.

```{r main-figure, fig.width=8, fig.height=5.5}
WATER_OFFSET <- 0.12
add_plot_x <- function(x) {
  x %>% mutate(x_plot = sqrt(ERW_level) +
                 if_else(Water_type == "Ground", -WATER_OFFSET, WATER_OFFSET))
}

doc_plot_data <- doc_data %>% filter(!is.na(DOC_mg_C_L)) %>% add_plot_x()
doc_plot_summary <- doc_summary %>% add_plot_x()

p_doc_main <- ggplot(doc_plot_data,
                     aes(x_plot, DOC_mg_C_L, colour = Water_type,
                         shape = Water_type)) +
  geom_point(position = position_jitter(width = 0.05, height = 0, seed = 123),
             size = 2, alpha = 0.45) +
  geom_line(data = doc_plot_summary,
            aes(y = mean, group = Water_type), linewidth = 0.8) +
  geom_errorbar(data = doc_plot_summary,
                aes(y = mean, ymin = pmax(mean - sd, 0), ymax = mean + sd),
                width = 0.10, linewidth = 0.7, na.rm = TRUE) +
  geom_point(data = doc_plot_summary, aes(y = mean), size = 3.8) +
  scale_x_continuous(breaks = sqrt(ERW_LEVELS), labels = ERW_LEVELS,
                     minor_breaks = NULL,
                     expand = expansion(mult = c(0.04, 0.07))) +
  scale_colour_manual(values = WATER_COLOURS, labels = WATER_LABELS) +
  scale_shape_manual(values = WATER_SHAPES, labels = WATER_LABELS) +
  labs(x = expression("ERW application (" * t ~ ha^{-1} * ")"),
       y = expression("DOC concentration (mg C L"^{-1}*")"),
       colour = "Water type", shape = "Water type",
       title = "Dissolved organic carbon",
       subtitle = "Raw observations and arithmetic mean ± SD") +
  DOC_THEME

p_doc_main
```

# 4. Model data and control audit

```{r model-data}
doc_model_data <- doc_data %>%
  filter(
    !is.na(DOC_mg_C_L),
    !is.na(ERW_level),
    !is.na(Water_type),
    !is.na(Block)
  ) %>%
  mutate(erw_raw_c = ERW_level - mean(ERW_level),
         erw_log_c = ERW_log - mean(ERW_log)) %>%
  droplevels()

doc_model_counts <- doc_model_data %>%
  count(Water_type, ERW_level, name = "n") %>%
  complete(Water_type, ERW_level = ERW_LEVELS, fill = list(n = 0))

if (!0 %in% doc_model_data$ERW_level) {
  stop("Zero-ERW controls are absent from doc_model_data.")
}
if (!all(ERW_LEVELS %in% doc_model_data$ERW_level)) {
  warning("At least one expected ERW level is absent from doc_model_data.")
}

doc_model_counts %>% kable(caption = "Observations in all DOC candidate models")
```

# 5. Stage 1: compare raw and log1p ERW

The two ML models below use the same response, observations, additive fixed effects and random effect. They are non-nested, so their functional forms are compared by AIC rather than a likelihood-ratio test. If their AIC values differ by less than 2, raw ERW is selected for interpretability; selection is not based on the ERW p-value.

```{r stage1}
fit_doc <- function(formula, reml = FALSE) {
  lmerTest::lmer(formula, data = doc_model_data, REML = reml,
                 na.action = na.exclude,
                 control = lme4::lmerControl(
                   optimizer = "bobyqa", optCtrl = list(maxfun = 200000)))
}

doc_raw_additive_ml <- fit_doc(
  DOC_mg_C_L ~ Water_type + erw_raw_c + (1 | Block))
doc_log_additive_ml <- fit_doc(
  DOC_mg_C_L ~ Water_type + erw_log_c + (1 | Block))

doc_erw_form_comparison <- tibble(
  erw_form = c("Raw ERW", "log1p(ERW)"),
  predictor = c("erw_raw_c", "erw_log_c"),
  AIC = c(AIC(doc_raw_additive_ml), AIC(doc_log_additive_ml)),
  BIC = c(BIC(doc_raw_additive_ml), BIC(doc_log_additive_ml)),
  logLik = c(as.numeric(logLik(doc_raw_additive_ml)),
             as.numeric(logLik(doc_log_additive_ml))),
  n = c(nobs(doc_raw_additive_ml), nobs(doc_log_additive_ml))
) %>% mutate(delta_AIC = AIC - min(AIC))

doc_erw_candidate_fixed <- bind_rows(
  broom.mixed::tidy(doc_raw_additive_ml, effects = "fixed", conf.int = TRUE) %>%
    mutate(erw_form = "Raw ERW", .before = 1),
  broom.mixed::tidy(doc_log_additive_ml, effects = "fixed", conf.int = TRUE) %>%
    mutate(erw_form = "log1p(ERW)", .before = 1)
)

tidy_type2 <- function(model, label) {
  as.data.frame(anova(model, type = 2, ddf = "Satterthwaite")) %>%
    rownames_to_column("term") %>% clean_names() %>%
    mutate(erw_form = label, .before = 1)
}
doc_erw_candidate_type2 <- bind_rows(
  tidy_type2(doc_raw_additive_ml, "Raw ERW"),
  tidy_type2(doc_log_additive_ml, "log1p(ERW)"))

if (AIC(doc_raw_additive_ml) <=
    AIC(doc_log_additive_ml) + AIC_TIE_THRESHOLD) {
  selected_erw_form <- "Raw ERW"
  selected_erw_var <- "erw_raw_c"
  doc_selected_additive_ml <- doc_raw_additive_ml
} else {
  selected_erw_form <- "log1p(ERW)"
  selected_erw_var <- "erw_log_c"
  doc_selected_additive_ml <- doc_log_additive_ml
}

doc_erw_form_comparison %>% arrange(AIC) %>%
  kable(digits = 3, caption = "Stage 1: ERW functional-form comparison")
selected_erw_form
```

# 6. Stage 2: test the interaction

```{r stage2}
doc_interaction_formula <- as.formula(paste0(
  "DOC_mg_C_L ~ Water_type * ", selected_erw_var, " + (1 | Block)"))
doc_selected_interaction_ml <- fit_doc(doc_interaction_formula)

doc_interaction_lrt <- anova(doc_selected_additive_ml,
                             doc_selected_interaction_ml, refit = FALSE)
doc_interaction_aic <- tibble(
  model = c("Additive", "Water type × ERW interaction"),
  selected_erw_form = selected_erw_form,
  AIC = c(AIC(doc_selected_additive_ml),
          AIC(doc_selected_interaction_ml)),
  BIC = c(BIC(doc_selected_additive_ml),
          BIC(doc_selected_interaction_ml)),
  logLik = c(as.numeric(logLik(doc_selected_additive_ml)),
             as.numeric(logLik(doc_selected_interaction_ml)))
) %>% mutate(delta_AIC = AIC - min(AIC))

interaction_p_value <- doc_interaction_lrt$`Pr(>Chisq)`[2]
if (!is.na(interaction_p_value) &&
    interaction_p_value < INTERACTION_ALPHA) {

  selected_structure <- "Interaction"

  doc_final_model <- fit_doc(
    stats::formula(doc_selected_interaction_ml),
    reml = TRUE
  )

} else {

  selected_structure <- "Additive"

  doc_final_model <- fit_doc(
    stats::formula(doc_selected_additive_ml),
    reml = TRUE
  )
}

doc_interaction_aic %>%
  kable(digits = 3, caption = "Stage 2: interaction comparison")
doc_interaction_lrt
selected_erw_form
selected_structure
summary(doc_final_model)
```

# 7. Final results

```{r final-results}
doc_fixed_effects <- broom.mixed::tidy(
  doc_final_model, effects = "fixed", conf.int = TRUE)
doc_type2_tests <- as.data.frame(
  anova(doc_final_model, type = 2, ddf = "Satterthwaite")) %>%
  rownames_to_column("term") %>% clean_names()
doc_random_effects <- broom.mixed::tidy(
  doc_final_model, effects = "ran_pars")
doc_model_audit <- tibble(
  selected_erw_form, selected_structure, interaction_p_value,
  n = nobs(doc_final_model),
  zero_erw_n = sum(doc_model_data$ERW_level == 0),
  singular_fit = lme4::isSingular(doc_final_model, tol = 1e-4))

doc_fixed_effects %>%
  kable(digits = 4, caption = paste("Final fixed effects:",
                                    selected_erw_form, selected_structure))
doc_type2_tests %>% kable(digits = 4, caption = "Final Type II tests")
doc_model_audit %>% kable(digits = 4, caption = "Final-model audit")
performance::r2_nakagawa(doc_final_model)
```

```{r trends-contrasts}
if (selected_structure == "Interaction") {
  doc_erw_trends <- emmeans::emtrends(
    doc_final_model, ~ Water_type, var = selected_erw_var)
  doc_erw_trend_contrast <- pairs(doc_erw_trends, adjust = "none")
} else {
  doc_erw_trends <- emmeans::emtrends(
    doc_final_model, ~ 1, var = selected_erw_var)
  doc_erw_trend_contrast <- NULL
}

doc_water_emmeans <- emmeans::emmeans(doc_final_model, ~ Water_type)
doc_water_contrast <- pairs(doc_water_emmeans, adjust = "none")
doc_erw_trends_table <- as.data.frame(
  summary(doc_erw_trends, infer = c(TRUE, TRUE))) %>% clean_names()
doc_erw_trend_contrast_table <- if (is.null(doc_erw_trend_contrast)) tibble() else
  as.data.frame(summary(doc_erw_trend_contrast,
                        infer = c(TRUE, TRUE))) %>% clean_names()
doc_water_emmeans_table <- as.data.frame(
  summary(doc_water_emmeans, infer = c(TRUE, TRUE))) %>% clean_names()
doc_water_contrast_table <- as.data.frame(
  summary(doc_water_contrast, infer = c(TRUE, TRUE))) %>% clean_names()

doc_erw_trends_table %>%
  kable(digits = 4, caption = paste("Estimated DOC trend using", selected_erw_form))
doc_water_contrast_table %>%
  kable(digits = 4, caption = "Adjusted DOC contrast between water types")
```

# 8. Diagnostics

```{r diagnostics, fig.width=9, fig.height=8}
old_par <- par(no.readonly = TRUE)
on.exit(par(old_par), add = TRUE)
par(mfrow = c(2, 2))
doc_residuals <- resid(doc_final_model, type = "pearson")
plot(fitted(doc_final_model), doc_residuals,
     xlab = "Fitted values", ylab = "Pearson residuals",
     main = "Residuals vs fitted")
abline(h = 0, lty = 2)
qqnorm(doc_residuals, main = "Normal Q–Q plot"); qqline(doc_residuals)
hist(doc_residuals, main = "Residual distribution", xlab = "Pearson residual")
plot(fitted(doc_final_model), sqrt(abs(doc_residuals)),
     xlab = "Fitted values", ylab = expression(sqrt("|Pearson residual|")),
     main = "Scale-location")
```

```{r diagnostic-statistics}
performance::check_collinearity(doc_final_model)
```

# 9. Export

```{r export}
ggsave(file.path(params$output_dir, "DOC_main_figure.png"), p_doc_main,
       width = 8, height = 5.5, dpi = 300)

doc_interaction_lrt_table <- as.data.frame(doc_interaction_lrt) %>%
  rownames_to_column("model") %>% clean_names()

write_xlsx(
  list(
    cleaned_data = doc_data %>% mutate(across(where(is.factor), as.character)),
    descriptive = doc_summary %>% mutate(Water_type = as.character(Water_type)),
    model_ERW_counts = doc_model_counts %>%
      mutate(Water_type = as.character(Water_type)),
    ERW_form_AIC = doc_erw_form_comparison,
    ERW_form_fixed = doc_erw_candidate_fixed,
    ERW_form_type2 = doc_erw_candidate_type2,
    interaction_AIC = doc_interaction_aic,
    interaction_LRT = doc_interaction_lrt_table,
    final_audit = doc_model_audit,
    final_fixed = doc_fixed_effects,
    final_type2 = doc_type2_tests,
    final_random = doc_random_effects,
    ERW_trends = doc_erw_trends_table,
    ERW_trend_contrast = doc_erw_trend_contrast_table,
    water_emmeans = doc_water_emmeans_table,
    water_contrast = doc_water_contrast_table
  ),
  file.path(params$output_dir, "DOC_analysis_tables.xlsx"))

saveRDS(doc_final_model,
        file.path(params$output_dir, "DOC_final_model.rds"))
```

# 10. Session information

```{r session-information}
sessionInfo()
```
