---
title: "Particulate Carbon Analysis: ERW Functional-Form Comparison"
author: "Xinlin Ni"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 7,
  dpi = 300
)
```

# 1. Analysis overview

This workflow analyses particulate organic carbon (POC) and particulate
inorganic carbon (PIC) in benthic and pelagic samples. All models include the
0 t ha^-1 controls. For each response, centred raw ERW and centred
`log1p(ERW)` are first compared in identical additive ML models. If
\(\Delta\mathrm{AIC}<2\), raw ERW is selected; otherwise, the lower-AIC form
is selected. The selected form is then used to test the
`Water_type × ERW` interaction, and the final model is refitted by REML.

The candidate overall models are:

```r
response_transformed ~ Water_type * erw_x +
  (1 | block / Mesocosm) + (1 | Date)

response_transformed ~ Water_type + erw_x +
  (1 | block / Mesocosm) + (1 | Date)
```

Date-specific supplementary models inherit the selected ERW form and use:

```r
response_transformed ~ Water_type + erw_x + (1 | block)
```

POC is modelled as `log(POC)` and PIC as `log1p(PIC)`. PIC equals total carbon
minus POC, with negative calculated values set to zero. The descriptive figure
uses square-root x-axis spacing only for display.

# 2. Packages

```{r packages}
library(tidyverse)
library(readxl)
library(janitor)
library(lme4)
library(lmerTest)
library(broom.mixed)
library(emmeans)
library(performance)
library(patchwork)
library(knitr)

options(contrasts = c("contr.treatment", "contr.poly"))
```

# 3. Settings

```{r settings}
BENTHIC_FILE <- "../data/benthic_carbon_mass.xlsx"
PELAGIC_FILE <- "../data/pelagic_carbon_mass.xlsx"

input_files <- c(
  BENTHIC_FILE,
  PELAGIC_FILE
)

if (any(!file.exists(input_files))) {
  stop(
    "Could not find: ",
    paste(
      input_files[!file.exists(input_files)],
      collapse = ", "
    )
  )
}

OUTPUT_DIR <- "../outputs/particulate_carbon"
FIGURE_DIR <- file.path(OUTPUT_DIR, "figures")

dir.create(
  OUTPUT_DIR,
  recursive = TRUE,
  showWarnings = FALSE
)

dir.create(
  FIGURE_DIR,
  recursive = TRUE,
  showWarnings = FALSE
)

FIGURE_DIR <- file.path(OUTPUT_DIR, "figures")

dir.create(OUTPUT_DIR, recursive = TRUE, showWarnings = FALSE)
dir.create(FIGURE_DIR, recursive = TRUE, showWarnings = FALSE)

ERW_LEVELS <- c(0, 5, 10, 20, 50, 100)

INTERACTION_ALPHA <- 0.05
AIC_EQUIVALENCE_THRESHOLD <- 2

WATER_COLOURS <- c(
  "Ground" = "#0072B2",
  "River" = "#D55E00"
)

WATER_LABELS <- c(
  "Ground" = "Groundwater",
  "River" = "River water"
)

WATER_SHAPES <- c(
  "Ground" = 16,
  "River" = 17
)

```

# 4. Experimental design

The same mesocosm design is used for both datasets. Mesocosms 1–6 belong to
block 1, 7–12 to block 2, and so on.

Pelagic samples collected on 19 May used 50 mL for every mesocosm. Samples
collected from 22 May onwards use the recorded mesocosm-specific filtration
volume.

```{r design}
design <- tribble(
  ~Mesocosm, ~Water_type, ~ERW_level, ~volume_from_22_may_ml,
  1,"Ground",5,240,    2,"Ground",50,240,   3,"Ground",10,300,
  4,"Ground",0,360,    5,"Ground",20,300,   6,"Ground",100,300,
  7,"River",10,300,    8,"River",20,300,    9,"River",0,240,
  10,"River",100,300, 11,"River",50,300,   12,"River",5,300,
  13,"River",100,300, 14,"River",0,240,    15,"River",5,300,
  16,"River",20,300,  17,"River",10,240,   18,"River",50,300,
  19,"Ground",20,300, 20,"Ground",100,300, 21,"Ground",5,240,
  22,"Ground",0,240,  23,"Ground",10,300,  24,"Ground",50,300,
  25,"Ground",0,300,  26,"Ground",50,300,  27,"Ground",10,300,
  28,"Ground",100,300,29,"Ground",20,300,  30,"Ground",5,300,
  31,"River",20,300,  32,"River",100,300,  33,"River",5,240,
  34,"River",50,300,  35,"River",0,300,    36,"River",10,240
) %>%
  mutate(
    Water_type = factor(
      Water_type,
      levels = c("Ground", "River")
    ),
    ERW_level = as.numeric(ERW_level),
    block = factor((Mesocosm - 1) %/% 6 + 1)
  )

stopifnot(
  nrow(design) == 36,
  setequal(design$ERW_level, ERW_LEVELS)
)
```

# 5. Data-processing functions

```{r processing-functions}
read_carbon_file <- function(path, compartment) {
  read_excel(path) %>%
    clean_names() %>%
    transmute(
      sample_name,
      Mesocosm = as.integer(mesocosm),
      sample_date = as.Date(date),
      carbon_percent = as.numeric(carbon_percent),
      mass_carbon_whole_filter_mg =
        as.numeric(mass_of_organic_carbon_in_whole_filters_mg),
      fraction = if_else(
        str_detect(str_to_upper(sample_name), "POC"),
        "POC",
        "TC"
      ),
      compartment = compartment,
      sample_status = if_else(
        is.na(carbon_percent) |
          is.na(mass_carbon_whole_filter_mg),
        "missing_or_not_measured",
        "measured"
      )
    )
}

prepare_carbon_data <- function(raw_data, compartment) {
  measured <- raw_data %>%
    filter(sample_status == "measured") %>%
    left_join(
      design,
      by = c("Mesocosm" = "Mesocosm")
    ) %>%
    mutate(
      Mesocosm = factor(Mesocosm),
      volume_ml = case_when(
        compartment == "Pelagic" &
          sample_date == as.Date("2025-05-19") ~ 50,
        TRUE ~ volume_from_22_may_ml
      ),
      conc_mg_L =
        mass_carbon_whole_filter_mg / volume_ml * 1000
    )

  wide <- measured %>%
    select(
      compartment,
      Mesocosm,
      block,
      sample_date,
      Water_type,
      ERW_level,
      fraction,
      conc_mg_L
    ) %>%
    pivot_wider(
      names_from = fraction,
      values_from = conc_mg_L,
      names_prefix = "conc_"
    ) %>%
    mutate(
      POC_exceeds_TC = conc_POC > conc_TC,
      PIC_raw = conc_TC - conc_POC,
      PIC_conc = pmax(PIC_raw, 0),
      sample_date_f = factor(
        sample_date,
        levels = sort(unique(sample_date))
      ),
      Water_type = factor(
        Water_type,
        levels = c("Ground", "River")
      ),
      ERW_level = as.numeric(ERW_level),
      block = factor(block),
      Mesocosm = factor(Mesocosm)
    ) %>%
    arrange(
      sample_date,
      Water_type,
      ERW_level,
      Mesocosm
    )

  wide
}
```

# 6. Import and prepare data

```{r import-prepare}
message(
  "Reading benthic data from: ",
  normalizePath(BENTHIC_FILE)
)

message(
  "Reading pelagic data from: ",
  normalizePath(PELAGIC_FILE)
)

benthic_raw <- read_carbon_file(
  BENTHIC_FILE,
  "Benthic"
)

pelagic_raw <- read_carbon_file(
  PELAGIC_FILE,
  "Pelagic"
)

benthic_wide <- prepare_carbon_data(
  benthic_raw,
  "Benthic"
)

pelagic_wide <- prepare_carbon_data(
  pelagic_raw,
  "Pelagic"
)

particulate_wide <- bind_rows(
  benthic_wide,
  pelagic_wide
)
```

# 7. Data checks

## 7.1 Join and block checks

```{r join-block-checks}
particulate_wide %>%
  summarise(
    missing_water_type = sum(is.na(Water_type)),
    missing_erw = sum(is.na(ERW_level)),
    missing_block = sum(is.na(block)),
    missing_mesocosm = sum(is.na(Mesocosm))
  )

particulate_wide %>%
  distinct(Mesocosm, block) %>%
  arrange(as.integer(as.character(Mesocosm)))
```

## 7.2 Missing or unmeasured samples

```{r missing-samples}
missing_samples <- bind_rows(
  benthic_raw,
  pelagic_raw
) %>%
  filter(sample_status != "measured") %>%
  arrange(
    compartment,
    sample_date,
    fraction,
    Mesocosm
  )

missing_samples
```

## 7.3 Completeness

```{r completeness}
completeness_table <- bind_rows(
  benthic_raw,
  pelagic_raw
) %>%
  count(
    compartment,
    sample_date,
    fraction,
    sample_status
  ) %>%
  pivot_wider(
    names_from = sample_status,
    values_from = n,
    values_fill = 0
  ) %>%
  arrange(
    compartment,
    sample_date,
    fraction
  )

kable(completeness_table)
```

## 7.4 PIC calculation checks

```{r pic-checks}
pic_checks <- particulate_wide %>%
  group_by(compartment, sample_date) %>%
  summarise(
    total_rows = n(),
    missing_POC = sum(is.na(conc_POC)),
    missing_TC = sum(is.na(conc_TC)),
    missing_PIC = sum(is.na(PIC_conc)),
    POC_greater_than_TC =
      sum(POC_exceeds_TC, na.rm = TRUE),
    zero_PIC_after_truncation =
      sum(PIC_conc == 0, na.rm = TRUE),
    .groups = "drop"
  )

kable(pic_checks)
```

```{r inspect-negative-pic}
particulate_wide %>%
  filter(POC_exceeds_TC %in% TRUE) %>%
  select(
    compartment,
    Mesocosm,
    block,
    sample_date,
    Water_type,
    ERW_level,
    conc_POC,
    conc_TC,
    PIC_raw,
    PIC_conc
  )
```

# 8. Descriptive summaries and plots

```{r descriptive-functions}
carbon_summary <- function(data, response) {
  data %>%
    filter(!is.na(.data[[response]])) %>%
    group_by(
      sample_date,
      Water_type,
      ERW_level
    ) %>%
    summarise(
      n = n(),
      mean = mean(
        .data[[response]],
        na.rm = TRUE
      ),
      sd = sd(
        .data[[response]],
        na.rm = TRUE
      ),
      .groups = "drop"
    )
}

plot_raw_mean_sd <- function(
    data,
    response,
    y_label,
    title
) {

  plot_data <- data %>%
    filter(!is.na(.data[[response]])) %>%
    mutate(
      x_plot = sqrt(ERW_level) +
        if_else(
          Water_type == "Ground",
          -0.10,
          0.10
        )
    )

  summary_data <- carbon_summary(
    data = data,
    response = response
  ) %>%
    mutate(
      x_plot = sqrt(ERW_level) +
        if_else(
          Water_type == "Ground",
          -0.10,
          0.10
        )
    )

  ggplot(
    data = plot_data,
    mapping = aes(
      x = x_plot,
      y = .data[[response]],
      colour = Water_type,
      shape = Water_type,
      group = Water_type
    )
  ) +

    geom_point(
      position = position_jitter(
        width = 0.035,
        height = 0,
        seed = 123
      ),
      size = 2,
      alpha = 0.45
    ) +

    geom_line(
      data = summary_data,
      mapping = aes(
        x = x_plot,
        y = mean,
        colour = Water_type,
        group = Water_type
      ),
      linewidth = 0.8,
      alpha = 0.9
    ) +

    geom_point(
      data = summary_data,
      mapping = aes(
        x = x_plot,
        y = mean
      ),
      size = 3,
      stroke = 0.8
    ) +

    geom_errorbar(
      data = summary_data,
      mapping = aes(
        x = x_plot,
        y = mean,
        ymin = pmax(mean - sd, 0),
        ymax = mean + sd
      ),
      width = 0.12,
      linewidth = 0.7,
      na.rm = TRUE
    ) +

    facet_wrap(
      ~ sample_date,
      nrow = 1,
      labeller = labeller(
        sample_date = function(x) {
          format(
            as.Date(x),
            "%Y-%m-%d"
          )
        }
      )
    ) +

    scale_x_continuous(
      breaks = sqrt(ERW_LEVELS),
      labels = ERW_LEVELS,
      minor_breaks = NULL,
      expand = expansion(
        mult = c(0.04, 0.07)
      )
    ) +

    scale_colour_manual(
      values = WATER_COLOURS,
      labels = WATER_LABELS
    ) +

    scale_shape_manual(
      values = WATER_SHAPES,
      labels = WATER_LABELS
    ) +

    labs(
      x = expression("ERW application (" * t ~ ha^{-1} * ")"),
      y = y_label,
      colour = "Water type",
      shape = "Water type",
      title = title,
      subtitle = paste(
        "Small points: raw observations;",
        "connected large points and error bars: mean ± SD"
      )
    ) +

    theme_classic() +

    theme(
      legend.position = "right",
      legend.direction = "vertical",
      strip.background = element_rect(
        fill = "grey95",
        colour = "grey70",
        linewidth = 0.5
      ),
      strip.text = element_text(
        face = "bold"
      )
    )
}

```

## 8.1 Combined benthic and pelagic figure

```{r benthic-plots, fig.width=11, fig.height=10}
p_benthic_poc <- plot_raw_mean_sd(
  benthic_wide,
  "conc_POC",
  expression(
    "POC (mg C L"^{-1}*")"
  ),
  "Benthic particulate organic carbon"
)

p_benthic_pic <- plot_raw_mean_sd(
  benthic_wide,
  "PIC_conc",
  expression(
    "PIC (mg C L"^{-1}*")"
  ),
  "Benthic particulate inorganic carbon"
)

large_text_theme <- theme(
  axis.title = element_text(size = 16, face = "bold"),
  axis.text = element_text(size = 14),
  strip.text = element_text(size = 14, face = "bold"),
  legend.title = element_text(size = 14),
  legend.text = element_text(size = 14)
)

p_benthic_poc <- p_benthic_poc + large_text_theme
p_benthic_pic <- p_benthic_pic + large_text_theme

p_benthic_combined <-
  (
    (p_benthic_poc +
       labs(title = NULL, subtitle = NULL, x = NULL) +
       theme(axis.title.x = element_blank())) /
    
    (p_benthic_pic +
       labs(title = NULL, subtitle = NULL, x = NULL) +
       theme(
         axis.title.x = element_blank(),
         strip.text.x = element_blank(),
         strip.background = element_blank()
       ))
  ) +
   plot_layout(guides = "collect") +
  plot_annotation(
    title = "(a)",
    theme = theme(
      plot.title = element_text(
        hjust = 0.05,
        size = 16,
        face = "bold"
      )
    )
  ) &
  theme(legend.position = "none")
```

```{r pelagic-plots, fig.width=24, fig.height=10}
p_pelagic_poc <- plot_raw_mean_sd(
  pelagic_wide,
  "conc_POC",
  expression(
    "Pelagic POC (mg C L"^{-1}*")"
  ),
  "Pelagic particulate organic carbon"
)

p_pelagic_pic <- plot_raw_mean_sd(
  pelagic_wide,
  "PIC_conc",
  expression(
    "Pelagic PIC (mg C L"^{-1}*")"
  ),
  "Pelagic particulate inorganic carbon"
)


p_pelagic_poc <- p_pelagic_poc + large_text_theme
p_pelagic_pic <- p_pelagic_pic + large_text_theme

p_pelagic_combined <-
  (
    (p_pelagic_poc +
       labs(title = NULL, subtitle = NULL, x = NULL, y = NULL) +
       theme(
         axis.title.x = element_blank(),
         axis.title.y = element_blank()
       )) /
    
    (p_pelagic_pic +
       labs(title = NULL, subtitle = NULL, x = NULL, y = NULL) +
       theme(
         axis.title.x = element_blank(),
         axis.title.y = element_blank(),
         strip.text.x = element_blank(),
         strip.background = element_blank()
       ))
  ) +
  plot_layout(guides = "collect") +
  plot_annotation(
    title = "(b)",
    theme = theme(
      plot.title = element_text(
        hjust = 0.03,
        size = 16,
        face = "bold"
      )
    )
  ) &
  theme(legend.position = "right")

p_particulate_mean_sd_combined <-
  (
    wrap_elements(full = p_benthic_combined) |
    wrap_elements(full = p_pelagic_combined)
  ) +
  plot_layout(widths = c(1, 1.17)) +
  plot_annotation(
    caption = expression("ERW application (" * t ~ ha^{-1} * ")"),
    theme = theme(
      plot.caption = element_text(
        hjust = 0.5,
        size = 16,
        margin = margin(t = 8)
      )
    )
  )

p_particulate_mean_sd_combined
```

## 8.2 Control means

```{r control-means}
control_carbon_means <- bind_rows(
  benthic_wide,
  pelagic_wide
) %>%
  filter(ERW_level == 0) %>%
  summarise(
    mean_control_POC = mean(conc_POC, na.rm = TRUE),
    mean_control_PIC = mean(PIC_conc, na.rm = TRUE)
  )

control_carbon_means
```

# 9. Mixed-model functions

```{r model-functions}
prepare_model_data <- function(
    data,
    response,
    transform = c("log", "log1p")
) {
  transform <- match.arg(transform)

  model_data <- data %>%
    filter(!is.na(.data[[response]]))

  if (
    transform == "log" &&
    any(model_data[[response]] <= 0, na.rm = TRUE)
  ) {
    stop(
      response,
      " contains zero or negative values and cannot use log()."
    )
  }

  model_data %>%
    mutate(
      response_transformed = case_when(
        transform == "log" ~
          log(.data[[response]]),
        transform == "log1p" ~
          log1p(.data[[response]])
      ),
      Water_type = droplevels(Water_type),
      sample_date_f = droplevels(sample_date_f),
      block = droplevels(block),
      Mesocosm = droplevels(Mesocosm)
    )
}

fit_overall_models <- function(
    data,
    response,
    transform,
    compartment,
    erw_form = c("raw", "log1p")
) {
  erw_form <- match.arg(erw_form)

  dat <- prepare_model_data(
    data,
    response,
    transform
  ) %>%
    mutate(
      erw_x = if (erw_form == "raw") {
        ERW_level - mean(ERW_level, na.rm = TRUE)
      } else {
        log1p(ERW_level) -
          mean(log1p(ERW_level), na.rm = TRUE)
      }
    )

  full_ml <- lmer(
    response_transformed ~
      Water_type * erw_x +
      (1 | block / Mesocosm) +
      (1 | sample_date_f),
    data = dat,
    REML = FALSE,
    na.action = na.exclude,
    control = lmerControl(
      optimizer = "bobyqa"
    )
  )

  additive_ml <- lmer(
    response_transformed ~
      Water_type +
      erw_x +
      (1 | block / Mesocosm) +
      (1 | sample_date_f),
    data = dat,
    REML = FALSE,
    na.action = na.exclude,
    control = lmerControl(
      optimizer = "bobyqa"
    )
  )

  comparison <- anova(
    additive_ml,
    full_ml
  )

  interaction_p <-
    comparison$`Pr(>Chisq)`[2]

  selected_structure <- if (
    is.na(interaction_p) ||
      interaction_p >= INTERACTION_ALPHA
  ) {
    "additive"
  } else {
    "interaction"
  }

  selected_ml <- if (
    selected_structure == "additive"
  ) {
    additive_ml
  } else {
    full_ml
  }

  final_model <- update(
    selected_ml,
    REML = TRUE
  )

  list(
    compartment = compartment,
    response = response,
    transform = transform,
    erw_form = erw_form,
    data = dat,
    full_ml = full_ml,
    additive_ml = additive_ml,
    comparison = comparison,
    interaction_p = interaction_p,
    selected_structure = selected_structure,
    selected_ml = selected_ml,
    final_model = final_model
  )
}

fit_daily_models <- function(
    data,
    response,
    transform,
    compartment,
    erw_form = c("raw", "log1p")
) {
  erw_form <- match.arg(erw_form)

  dat <- prepare_model_data(
    data,
    response,
    transform
  ) %>%
    mutate(
      erw_x = if (erw_form == "raw") {
        ERW_level - mean(ERW_level, na.rm = TRUE)
      } else {
        log1p(ERW_level) -
          mean(log1p(ERW_level), na.rm = TRUE)
      }
    )

  dat %>%
    group_by(
      sample_date,
      sample_date_f
    ) %>%
    nest() %>%
    mutate(
      compartment = compartment,
      response = response,
      transform = transform,
      erw_form = erw_form,
      selected_structure = "additive",
      model = map(
        data,
        ~ lmer(
            response_transformed ~
              Water_type +
              erw_x +
              (1 | block),
            data = .x,
            REML = TRUE,
            na.action = na.exclude,
            control = lmerControl(
              optimizer = "bobyqa"
            )
          )
      )
    ) %>%
    ungroup()
}

safe_type2_anova <- function(model) {
  anova(
    model,
    type = 2,
    ddf = "Satterthwaite"
  ) %>%
    as.data.frame() %>%
    rownames_to_column("term") %>%
    clean_names()
}

extract_overall_outputs <- function(fit) {
  final_model <- fit$final_model

  list(
    comparison =
      as.data.frame(fit$comparison) %>%
      rownames_to_column("candidate_model") %>%
      clean_names() %>%
      mutate(
        compartment = fit$compartment,
        response = fit$response,
        erw_form = fit$erw_form,
        selected_structure =
          fit$selected_structure,
        interaction_p =
          fit$interaction_p,
        .before = 1
      ),

    type2 =
      safe_type2_anova(final_model) %>%
      mutate(
        compartment = fit$compartment,
        response = fit$response,
        erw_form = fit$erw_form,
        selected_structure =
          fit$selected_structure,
        .before = 1
      ),

    fixed =
      broom.mixed::tidy(
        final_model,
        effects = "fixed",
        conf.int = TRUE
      ) %>%
      mutate(
        compartment = fit$compartment,
        response = fit$response,
        erw_form = fit$erw_form,
        selected_structure =
          fit$selected_structure,
        .before = 1
      ),

    random =
      broom.mixed::tidy(
        final_model,
        effects = "ran_pars",
        conf.int = TRUE
      ) %>%
      mutate(
        compartment = fit$compartment,
        response = fit$response,
        erw_form = fit$erw_form,
        selected_structure =
          fit$selected_structure,
        .before = 1
      ),

    fit =
      broom.mixed::glance(final_model) %>%
      mutate(
        compartment = fit$compartment,
        response = fit$response,
        erw_form = fit$erw_form,
        selected_structure =
          fit$selected_structure,
        interaction_p =
          fit$interaction_p,
        singular =
          performance::check_singularity(
            final_model
          ),
        .before = 1
      )
  )
}

extract_daily_outputs <- function(daily_models) {
  model_fit <- daily_models %>%
    transmute(
      compartment,
      response,
      erw_form,
      sample_date,
      sample_date_f,
      selected_structure,
      model
    ) %>%
    mutate(
      model_fit = map(
        model,
        broom.mixed::glance
      )
    ) %>%
    select(-model) %>%
    unnest(model_fit)

  coefficients <- daily_models %>%
    transmute(
      compartment,
      response,
      erw_form,
      sample_date,
      sample_date_f,
      selected_structure,
      model
    ) %>%
    mutate(
      coefficients = map(
        model,
        ~ broom.mixed::tidy(
            .x,
            effects = "fixed",
            conf.int = TRUE
          )
      )
    ) %>%
    select(-model) %>%
    unnest(coefficients)

  term_tests <- daily_models %>%
    transmute(
      compartment,
      response,
      erw_form,
      sample_date,
      sample_date_f,
      selected_structure,
      model
    ) %>%
    mutate(
      term_tests = map(
        model,
        safe_type2_anova
      )
    ) %>%
    select(-model) %>%
    unnest(term_tests) %>%
    group_by(
      compartment,
      response,
      erw_form,
      term
    ) %>%
    mutate(
      p_value_fdr =
        p.adjust(pr_f, method = "BH")
    ) %>%
    ungroup()

  erw_coefficients <- coefficients %>%
    filter(
      term == "erw_x"
    ) %>%
    group_by(
      compartment,
      response,
      erw_form,
      term
    ) %>%
    mutate(
      p_value_fdr =
        p.adjust(.data$p.value, method = "BH")
    ) %>%
    ungroup()

  checks <- daily_models %>%
    transmute(
      compartment,
      response,
      erw_form,
      sample_date,
      sample_date_f,
      selected_structure,
      singular = map_lgl(
        model,
        performance::check_singularity
      )
    )

  list(
    model_fit = model_fit,
    coefficients = coefficients,
    term_tests = term_tests,
    erw_coefficients = erw_coefficients,
    checks = checks
  )
}
```

# 10. Select ERW form, then test the interaction

For each response, raw ERW and `log1p(ERW)` are first compared using identical
additive models. Both forms use exactly the same observations and response
transformation, and both include the zero-ERW controls. After the ERW form has
been selected, the additive and water-type interaction structures are compared
only for that selected form.

```{r erw-form-selection-functions}
fit_additive_erw_form_ml <- function(
    data,
    response,
    transform,
    compartment,
    erw_form = c("raw", "log1p")
) {
  erw_form <- match.arg(erw_form)

  dat <- prepare_model_data(
    data,
    response,
    transform
  ) %>%
    mutate(
      erw_x = if (erw_form == "raw") {
        ERW_level - mean(ERW_level, na.rm = TRUE)
      } else {
        log1p(ERW_level) -
          mean(log1p(ERW_level), na.rm = TRUE)
      }
    )

  model <- lmer(
    response_transformed ~
      Water_type +
      erw_x +
      (1 | block / Mesocosm) +
      (1 | sample_date_f),
    data = dat,
    REML = FALSE,
    na.action = na.exclude,
    control = lmerControl(
      optimizer = "bobyqa"
    )
  )

  list(
    compartment = compartment,
    response = response,
    transform = transform,
    erw_form = erw_form,
    data = dat,
    model = model
  )
}

fit_both_erw_forms <- function(
    data,
    response,
    transform,
    compartment
) {
  raw_candidate <- fit_additive_erw_form_ml(
    data = data,
    response = response,
    transform = transform,
    compartment = compartment,
    erw_form = "raw"
  )

  log1p_candidate <- fit_additive_erw_form_ml(
    data = data,
    response = response,
    transform = transform,
    compartment = compartment,
    erw_form = "log1p"
  )

  candidates <- list(
    raw = raw_candidate,
    log1p = log1p_candidate
  )

  comparison <- imap_dfr(
    candidates,
    function(fit, form_name) {
      tibble(
        compartment = fit$compartment,
        response = fit$response,
        response_transform = fit$transform,
        erw_form = form_name,
        fixed_structure = "Water_type + ERW",
        n = nobs(fit$model),
        logLik = as.numeric(logLik(fit$model)),
        AIC = AIC(fit$model),
        BIC = BIC(fit$model)
      )
    }
  ) %>%
    arrange(AIC) %>%
    mutate(delta_AIC = AIC - min(AIC))

  aic_difference <- diff(range(comparison$AIC))

  selected_form <- if (
    aic_difference < AIC_EQUIVALENCE_THRESHOLD
  ) {
    "raw"
  } else {
    comparison$erw_form[which.min(comparison$AIC)]
  }

  comparison <- comparison %>%
    mutate(
      selected_final = erw_form == selected_form,
      selection_rule = if_else(
        aic_difference < AIC_EQUIVALENCE_THRESHOLD,
        "Delta AIC < 2: raw ERW preferred for simpler interpretation",
        "Delta AIC >= 2: lowest-AIC ERW form selected"
      )
    )

  # Only after selecting the ERW form do we test the interaction.
  selected_fit <- fit_overall_models(
    data = data,
    response = response,
    transform = transform,
    compartment = compartment,
    erw_form = selected_form
  )

  list(
    candidates = candidates,
    comparison = comparison,
    selected_form = selected_form,
    selected_fit = selected_fit
  )
}
```

```{r overall-models}
forms_benthic_poc <- fit_both_erw_forms(
  benthic_wide,
  response = "conc_POC",
  transform = "log",
  compartment = "Benthic"
)

forms_benthic_pic <- fit_both_erw_forms(
  benthic_wide,
  response = "PIC_conc",
  transform = "log1p",
  compartment = "Benthic"
)

forms_pelagic_poc <- fit_both_erw_forms(
  pelagic_wide,
  response = "conc_POC",
  transform = "log",
  compartment = "Pelagic"
)

forms_pelagic_pic <- fit_both_erw_forms(
  pelagic_wide,
  response = "PIC_conc",
  transform = "log1p",
  compartment = "Pelagic"
)

erw_form_fit_sets <- list(
  benthic_POC = forms_benthic_poc,
  benthic_PIC = forms_benthic_pic,
  pelagic_POC = forms_pelagic_poc,
  pelagic_PIC = forms_pelagic_pic
)

erw_form_comparison <- imap_dfr(
  erw_form_fit_sets,
  ~ .x$comparison %>% mutate(outcome = .y, .before = 1)
)

erw_form_comparison

erw_form_candidate_fixed_effects <- imap_dfr(
  erw_form_fit_sets,
  function(form_set, outcome_name) {
    imap_dfr(
      form_set$candidates,
      function(candidate, form_name) {
        broom.mixed::tidy(
          candidate$model,
          effects = "fixed",
          conf.int = TRUE
        ) %>%
          mutate(
            outcome = outcome_name,
            compartment = candidate$compartment,
            response = candidate$response,
            response_transform = candidate$transform,
            erw_form = form_name,
            .before = 1
          )
      }
    )
  }
)

erw_form_candidate_type2_tests <- imap_dfr(
  erw_form_fit_sets,
  function(form_set, outcome_name) {
    imap_dfr(
      form_set$candidates,
      function(candidate, form_name) {
        safe_type2_anova(candidate$model) %>%
          mutate(
            outcome = outcome_name,
            compartment = candidate$compartment,
            response = candidate$response,
            response_transform = candidate$transform,
            erw_form = form_name,
            .before = 1
          )
      }
    )
  }
)

erw_form_candidate_fixed_effects
erw_form_candidate_type2_tests

# These selected fits are used by every downstream overall and daily analysis.
fit_benthic_poc <- forms_benthic_poc$selected_fit
fit_benthic_pic <- forms_benthic_pic$selected_fit
fit_pelagic_poc <- forms_pelagic_poc$selected_fit
fit_pelagic_pic <- forms_pelagic_pic$selected_fit

overall_fits <- list(
  benthic_POC = fit_benthic_poc,
  benthic_PIC = fit_benthic_pic,
  pelagic_POC = fit_pelagic_poc,
  pelagic_PIC = fit_pelagic_pic
)

selected_model_treatment_balance <- imap_dfr(
  overall_fits,
  ~ .x$data %>%
    count(ERW_level, name = "n") %>%
    mutate(outcome = .y, .before = 1)
)

if (
  any(
    !map_lgl(
      overall_fits,
      ~ any(.x$data$ERW_level == 0, na.rm = TRUE)
    )
  )
) {
  stop("A selected model dataset does not contain the 0 t ha^-1 control.")
}

selected_model_treatment_balance
```

## 10.1 Interaction-model comparisons

```{r overall-model-comparisons}
overall_model_comparison <- imap_dfr(
  overall_fits,
  ~ extract_overall_outputs(.x)$comparison
)

overall_model_comparison
```

## 10.2 Final model summaries

```{r final-model-summaries}
summary(
  fit_benthic_poc$final_model
)

summary(
  fit_benthic_pic$final_model
)

summary(
  fit_pelagic_poc$final_model
)

summary(
  fit_pelagic_pic$final_model
)
```

## 10.3 Type II tests

```{r overall-type2}
overall_type2_tests <- imap_dfr(
  overall_fits,
  ~ extract_overall_outputs(.x)$type2
)

overall_type2_tests
```

## 10.4 Fixed and random effects

```{r overall-effects}
overall_fixed_effects <- imap_dfr(
  overall_fits,
  ~ extract_overall_outputs(.x)$fixed
)

overall_random_effects <- imap_dfr(
  overall_fits,
  ~ extract_overall_outputs(.x)$random
)

overall_fixed_effects
overall_random_effects
```

## 10.5 Model-fit information

```{r overall-fit}
overall_model_fit <- imap_dfr(
  overall_fits,
  ~ extract_overall_outputs(.x)$fit
)

overall_model_fit
```

# 11. Estimated trends and predictions

```{r prediction-functions}
make_prediction_grid <- function(fit) {
  dat <- fit$data
  model <- fit$final_model

  newdata <- expand_grid(
    ERW_level = ERW_LEVELS,
    Water_type = levels(dat$Water_type),
    sample_date_f = levels(dat$sample_date_f)[1]
  ) %>%
    mutate(
      erw_x = if (fit$erw_form == "raw") {
        ERW_level - mean(dat$ERW_level, na.rm = TRUE)
      } else {
        log1p(ERW_level) -
          mean(log1p(dat$ERW_level), na.rm = TRUE)
      },
      Water_type = factor(
        Water_type,
        levels = levels(dat$Water_type)
      ),
      sample_date_f = factor(
        sample_date_f,
        levels = levels(dat$sample_date_f)
      )
    )

  pred <- predict(
    model,
    newdata = newdata,
    re.form = NA,
    se.fit = TRUE
  )

  newdata %>%
    mutate(
      estimate_link = as.numeric(pred$fit),
      se_link = as.numeric(pred$se.fit),
      conf_low_link =
        estimate_link - 1.96 * se_link,
      conf_high_link =
        estimate_link + 1.96 * se_link,
      estimate = if (
        fit$transform == "log"
      ) {
        exp(estimate_link)
      } else {
        expm1(estimate_link)
      },
      conf_low = if (
        fit$transform == "log"
      ) {
        exp(conf_low_link)
      } else {
        pmax(expm1(conf_low_link), 0)
      },
      conf_high = if (
        fit$transform == "log"
      ) {
        exp(conf_high_link)
      } else {
        pmax(expm1(conf_high_link), 0)
      },
      compartment = fit$compartment,
      response = fit$response,
      erw_form = fit$erw_form,
      selected_structure =
        fit$selected_structure,
      .before = 1
    )
}

overall_predictions <- imap_dfr(
  overall_fits,
  ~ make_prediction_grid(.x)
)

overall_predictions

pic_difference <- overall_predictions %>%
  filter(
    compartment == "Benthic",
    response == "PIC_conc",
    ERW_level %in% c(0, 100)
  ) %>%
  select(Water_type, ERW_level, estimate) %>%
  pivot_wider(
    names_from = ERW_level,
    values_from = estimate,
    names_prefix = "ERW_"
  ) %>%
  mutate(
    delta_PIC = ERW_100 - ERW_0,
    potential_CO2 = delta_PIC * 44.01 / 12.01
  )

pic_difference

pic_difference_overall <- pic_difference %>%
  summarise(
    delta_PIC = mean(delta_PIC),
    potential_CO2 = mean(potential_CO2)
  )

pic_difference_overall
```

```{r overall-trends}
overall_erw_slopes <- imap_dfr(
  overall_fits,
  function(fit, name) {
    trend <- emtrends(
      fit$final_model,
      specs = ~ Water_type,
      var = "erw_x"
    )

    as.data.frame(trend) %>%
      clean_names() %>%
      mutate(
        compartment = fit$compartment,
        response = fit$response,
        erw_form = fit$erw_form,
        selected_structure =
          fit$selected_structure,
        .before = 1
      )
  }
)

overall_erw_slopes
```

```{r water-emmeans}
water_type_emmeans <- imap_dfr(
  overall_fits,
  function(fit, name) {
    emm <- emmeans(
      fit$final_model,
      specs = ~ Water_type
    )

    as.data.frame(emm) %>%
      clean_names() %>%
      mutate(
        compartment = fit$compartment,
        response = fit$response,
        erw_form = fit$erw_form,
        selected_structure =
          fit$selected_structure,
        .before = 1
      )
  }
)

water_type_emmeans
```

# 12. Daily models

```{r daily-models}
daily_benthic_poc <- fit_daily_models(
  benthic_wide,
  response = "conc_POC",
  transform = "log",
  compartment = "Benthic",
  erw_form = fit_benthic_poc$erw_form
)

daily_benthic_pic <- fit_daily_models(
  benthic_wide,
  response = "PIC_conc",
  transform = "log1p",
  compartment = "Benthic",
  erw_form = fit_benthic_pic$erw_form
)

daily_pelagic_poc <- fit_daily_models(
  pelagic_wide,
  response = "conc_POC",
  transform = "log",
  compartment = "Pelagic",
  erw_form = fit_pelagic_poc$erw_form
)

daily_pelagic_pic <- fit_daily_models(
  pelagic_wide,
  response = "PIC_conc",
  transform = "log1p",
  compartment = "Pelagic",
  erw_form = fit_pelagic_pic$erw_form
)

all_daily_models <- bind_rows(
  daily_benthic_poc,
  daily_benthic_pic,
  daily_pelagic_poc,
  daily_pelagic_pic
)
```

## 12.1 Daily model outputs

```{r daily-outputs}
daily_outputs <- extract_daily_outputs(
  all_daily_models
)

daily_model_fit <-
  daily_outputs$model_fit

daily_model_coefficients <-
  daily_outputs$coefficients

daily_model_term_tests <-
  daily_outputs$term_tests

daily_erw_coefficients <-
  daily_outputs$erw_coefficients

daily_model_checks <-
  daily_outputs$checks

daily_model_fit
daily_model_term_tests
daily_erw_coefficients
daily_model_checks
```

# 13. Diagnostics

```{r overall-diagnostics-benthic-poc, fig.width=10, fig.height=8}
check_model(
  fit_benthic_poc$final_model
)
```

```{r overall-diagnostics-benthic-pic, fig.width=10, fig.height=8}
check_model(
  fit_benthic_pic$final_model
)
```

```{r overall-diagnostics-pelagic-poc, fig.width=10, fig.height=8}
check_model(
  fit_pelagic_poc$final_model
)
```

```{r overall-diagnostics-pelagic-pic, fig.width=10, fig.height=8}
check_model(
  fit_pelagic_pic$final_model
)
```

# 14. Export summary tables

```{r build-summary-table}
particulate_summary <- bind_rows(

  carbon_summary(
    benthic_wide,
    "conc_POC"
  ) %>%
    mutate(
      compartment = "Benthic",
      response = "POC"
    ),

  carbon_summary(
    benthic_wide,
    "PIC_conc"
  ) %>%
    mutate(
      compartment = "Benthic",
      response = "PIC"
    ),

  carbon_summary(
    pelagic_wide,
    "conc_POC"
  ) %>%
    mutate(
      compartment = "Pelagic",
      response = "POC"
    ),

  carbon_summary(
    pelagic_wide,
    "PIC_conc"
  ) %>%
    mutate(
      compartment = "Pelagic",
      response = "PIC"
    )

) %>%
  select(
    compartment,
    response,
    sample_date,
    Water_type,
    ERW_level,
    everything()
  ) %>%
  arrange(
    compartment,
    response,
    sample_date,
    Water_type,
    ERW_level
  )

particulate_summary
```

# 15. Export results

```{r export-csv}
write_csv(
  erw_form_comparison,
  file.path(OUTPUT_DIR, "particulate_ERW_form_comparison.csv")
)

write_csv(
  overall_model_comparison,
  file.path(OUTPUT_DIR, "particulate_interaction_comparison.csv")
)

write_csv(
  particulate_summary,
  file.path(OUTPUT_DIR, "particulate_summary_mean_sd.csv")
)

write_csv(
  overall_type2_tests,
  file.path(OUTPUT_DIR, "particulate_final_model_type2_tests.csv")
)

write_csv(
  overall_fixed_effects,
  file.path(OUTPUT_DIR, "particulate_final_model_fixed_effects.csv")
)

write_csv(
  overall_erw_slopes,
  file.path(OUTPUT_DIR, "particulate_overall_ERW_slopes.csv")
)

write_csv(
  overall_predictions,
  file.path(OUTPUT_DIR, "particulate_final_model_predictions.csv")
)

write_csv(
  daily_model_term_tests,
  file.path(OUTPUT_DIR, "particulate_daily_model_term_tests.csv")
)

write_csv(
  daily_erw_coefficients,
  file.path(OUTPUT_DIR, "particulate_daily_ERW_coefficients.csv")
)
```

# 17. Session information

```{r session-info}
sessionInfo()
```
