---
title: "Enhanced Rock Weathering Mesocosm Experiment: K600 Screening and Stream Metabolism"
author: "Xinlin Ni"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
params:
  data_file: "/Users/nixl/Documents/IC/dissertation/code/dissoved_oxygen/combined_do_light_data.csv"
  meso9_file: "/Users/nixl/Documents/IC/dissertation/code/dissoved_oxygen/Meso_9_clean.xlsx"
  output_dir: "/Users/nixl/Documents/IC/dissertation/code/version_2/metabolism/v_2/results"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 10,
  fig.height = 8,
  dpi = 300
)

if (!exists("params")) {
  params <- list(
    data_file  = "/Users/nixl/Documents/IC/dissertation/code/dissoved_oxygen/combined_do_light_data.csv",
    meso9_file = "/Users/nixl/Documents/IC/dissertation/code/dissoved_oxygen/Meso_9_clean.xlsx",
    output_dir = "/Users/nixl/Documents/IC/dissertation/code/version_2/metabolism/v_2/results"
  )
}
```

# 1. Packages and global settings

```{r load_libraries}
library(tidyverse)
library(readxl)
library(lubridate)
library(suncalc)
library(lme4)
library(lmerTest)
library(ggeffects)
library(broom.mixed)
library(MuMIn)
library(patchwork)
library(scales)
library(DHARMa)
library(knitr)

WATER_COLOURS    <- c("Ground" = "#0072B2", "River" = "#D55E00")
WATER_LABELS     <- c("Ground" = "Groundwater", "River" = "River water")
WATER_SHAPES     <- c("Ground" = 16, "River" = 17)
METRIC_COLOURS   <- c("GPP" = "#2196F3", "ER" = "#F44336")
TREATMENT_LEVELS <- c(0, 5, 10, 20, 50, 100)

if (!dir.exists(params$output_dir)) {
  dir.create(params$output_dir, recursive = TRUE)
}

theme_xinlin <- function(base_size = 11) {
  theme_classic(base_size = base_size) +
    theme(
      plot.title       = element_text(hjust = 0.5, face = "bold"),
      legend.position  = "top",
      strip.background = element_rect(fill = "grey95", colour = "grey70"),
      strip.text       = element_text(face = "bold"),
      axis.text.x      = element_text(angle = 45, hjust = 1)
    )
}
```

# 2. Read and inspect the data

```{r read_data}
if (!file.exists(params$data_file)) {
  stop("Data file not found: ", params$data_file)
}

meso9_file_candidates <- unique(c(
  params$meso9_file,
  "Meso_9_clean.xlsx",
  "Meso_9_clean(1).xlsx",
  file.path(dirname(params$data_file), "Meso_9_clean.xlsx"),
  file.path(dirname(params$data_file), "Meso_9_clean(1).xlsx")
))

meso9_file <- meso9_file_candidates[
  file.exists(meso9_file_candidates)
][1]

if (is.na(meso9_file)) {
  stop(
    "Corrected Mesocosm 9 file not found. ",
    "Place Meso_9_clean.xlsx beside this Rmd or update params$meso9_file."
  )
}

df_original <- read_csv(params$data_file) %>%
  mutate(timestamp = ymd_hms(timestamp, tz = "UTC"))

meso9_metadata <- df_original %>%
  filter(as.character(mesocosm) == "9") %>%
  distinct(mesocosm, treatment_label, water_type, treatment) %>%
  slice(1)

if (nrow(meso9_metadata) == 0) {
  meso9_metadata <- tibble(
    mesocosm = 9,
    treatment_label = "0 t/ha",
    water_type = "River",
    treatment = 0
  )
}

parse_meso9_timestamp <- function(x) {
  if (inherits(x, "POSIXt")) {
    return(with_tz(x, "UTC"))
  }

  if (inherits(x, "Date")) {
    return(as_datetime(x, tz = "UTC"))
  }

  if (is.numeric(x)) {
    return(as.POSIXct(
      x * 86400,
      origin = "1899-12-30",
      tz = "UTC"
    ))
  }

  ymd_hms(x, tz = "UTC")
}

meso9_corrected <- read_excel(
  path = meso9_file,
  sheet = "Meso_9_clean",
  range = cell_cols("A:F"),
  na = c("", "NA")
) %>%
  transmute(
    timestamp = parse_meso9_timestamp(TIMESTAMP),
    mesocosm = meso9_metadata$mesocosm[[1]],
    treatment_label = meso9_metadata$treatment_label[[1]],
    water_type = meso9_metadata$water_type[[1]],
    treatment = meso9_metadata$treatment[[1]],
    Temp = as.numeric(Tw_9_Avg),
    temp_c = as.numeric(Tw_9_Avg),
    do_corrected = as.numeric(DO_conc_final)
  )

df <- df_original %>%
  filter(as.character(mesocosm) != "9") %>%
  bind_rows(meso9_corrected) %>%
  filter(
    timestamp >= as.Date("2025-05-15"),
    timestamp < as.Date("2025-06-13")
  ) %>%
  arrange(mesocosm, timestamp) %>%
  mutate(temp_final = ifelse(is.na(Temp), temp_c, Temp))

glimpse(df)

cat(
  "Corrected Mesocosm 9 observations included:",
  sum(as.character(df$mesocosm) == "9"),
  "\n"
)

treatment_info <- df %>%
  distinct(mesocosm, treatment_label, water_type, treatment)

treatment_info %>%
  arrange(water_type, treatment) %>%
  kable(caption = "Mesocosm treatment allocation")
```

# 3. Global parameters

```{r global_params}
depth <- 0.15
bpabs <- 760
dt    <- 15 / (60 * 24)
```

# 4. Metabolism calculation functions

```{r helper_functions}
# O2 saturation (Garcia & Gordon 1992, Eq. 34.16)
calc_osat <- function(temp, bp = 760) {
  ts <- log((298.15 - temp) / (273.15 + temp))
  a0 <- 2.00907; a1 <- 3.22014; a2 <- 4.0501
  a3 <- 4.94457; a4 <- -0.256847; a5 <- 3.88767
  u  <- 10^(8.10765 - 1750.3 / (235 + temp))
  osat_std <- exp(a0 + a1*ts + a2*ts^2 + a3*ts^3 + a4*ts^4 + a5*ts^5)
  osat_std * (bp - u) / (760 - u) * 1.42905
}

# K600 -> KO (Wanninkhof 1992, Eq. 34.15)
calc_KO <- function(temp, K600) {
  K600 / ((600 / (1800.6 - 120.10*temp + 3.7818*temp^2 - 0.047608*temp^3))^(-0.5))
}

# Sunrise/sunset (Birmingham: 52.45N, -1.93W)
get_sun_times <- function(date, lat = 52.45, lon = -1.93) {
  st <- getSunlightTimes(
    date = date, lat = lat, lon = lon,
    keep = c("sunrise", "sunset"), tz = "UTC"
  )
  list(sunrise = st$sunrise, sunset = st$sunset)
}

# Core metabolism calculation (Hall & Hotchkiss 2017, Section 34.3.1.5)
calc_metabolism_day <- function(data_day, K600, depth, bp = 760) {
  data_day <- data_day %>%
    arrange(timestamp) %>%
    mutate(
      osat         = calc_osat(temp_final, bp),
      KO           = calc_KO(temp_final, K600),
      deficit      = osat - do_corrected,
      gas_exchange = KO * deficit,
      dO_dt        = (lead(do_corrected) - do_corrected) / dt,
      metab_vol    = dO_dt - gas_exchange,
      metab_area   = metab_vol * .env$depth
    )

  the_date <- as.Date(data_day$timestamp[1])
  sun      <- get_sun_times(the_date)

  data_day <- data_day %>%
    mutate(is_night = timestamp < sun$sunrise | timestamp >= sun$sunset)

  ER <- data_day %>%
    filter(is_night, !is.na(metab_area)) %>%
    pull(metab_area) %>%
    mean(na.rm = TRUE)

  daytime <- data_day %>% filter(!is_night, !is.na(metab_area))
  if (nrow(daytime) == 0) return(tibble(GPP = NA, ER = ER))

  GPP <- sum((daytime$metab_area - ER) * dt, na.rm = TRUE)
  tibble(GPP = GPP, ER = ER)
}

# Estimate daily K600 from nighttime oxygen dynamics
nighttime_regression <- function(data_day, bp = 760) {
  data_day <- data_day %>%
    arrange(timestamp) %>%
    mutate(
      osat    = calc_osat(temp_final, bp),
      deficit = osat - do_corrected,
      dO_dt   = (lead(do_corrected) - do_corrected) / dt
    )

  the_date <- as.Date(data_day$timestamp[1])
  sun      <- get_sun_times(the_date)

  night_data <- data_day %>%
    filter(timestamp < sun$sunrise | timestamp >= sun$sunset) %>%
    filter(!is.na(dO_dt), !is.na(deficit))

  if (nrow(night_data) < 5) {
    return(tibble(KO_nightreg = NA_real_, K600_nightreg = NA_real_))
  }

  fit       <- lm(dO_dt ~ deficit, data = night_data)
  KO_est    <- unname(coef(fit)["deficit"])
  mean_temp <- mean(night_data$temp_final, na.rm = TRUE)
  ScO2      <- 1800.6 - 120.10 * mean_temp +
    3.7818 * mean_temp^2 - 0.047608 * mean_temp^3
  K600_est  <- KO_est * (600 / ScO2)^0.5

  tibble(KO_nightreg = KO_est, K600_nightreg = K600_est)
}
```

# 5. K600 method screening

First, GPP and ER are calculated using three K600 approaches: the fixed
literature value of 6.72, a fixed value equal to the median of valid nighttime
regression estimates, and daily dynamic nighttime-regression estimates. The
fixed literature value is retained for all subsequent figures and models.

## 5.1 Estimate K600 by nighttime regression

```{r estimate_nighttime_K600}
nightreg_results <- df %>%
  filter(!is.na(do_corrected), !is.na(temp_final)) %>%
  mutate(date = as.Date(timestamp)) %>%
  group_by(mesocosm, date) %>%
  group_modify(~ nighttime_regression(.x)) %>%
  ungroup()

nightreg_clean <- nightreg_results %>%
  filter(
    is.finite(K600_nightreg),
    K600_nightreg > 0,
    K600_nightreg < 50
  )

K600_median <- median(nightreg_clean$K600_nightreg, na.rm = TRUE)

K600_estimation_summary <- tibble(
  Total_mesocosm_days = nrow(nightreg_results),
  Retained_mesocosm_days = nrow(nightreg_clean),
  Fixed_literature_K600 = 6.72,
  Median_nighttime_regression_K600 = round(K600_median, 3)
)

K600_estimation_summary %>%
  kable(caption = "Summary of nighttime-regression K600 estimates")
```

## 5.2 Calculate GPP and ER using the three K600 methods

```{r calculate_three_K600_methods}
# Method A: fixed literature value
results_6.72 <- df %>%
  filter(!is.na(do_corrected), !is.na(temp_final)) %>%
  mutate(date = as.Date(timestamp)) %>%
  group_by(mesocosm, date) %>%
  group_modify(~ calc_metabolism_day(.x, K600 = 6.72, depth = depth)) %>%
  ungroup()

# Method B: fixed nighttime-regression median
results_nightreg_median <- df %>%
  filter(!is.na(do_corrected), !is.na(temp_final)) %>%
  mutate(date = as.Date(timestamp)) %>%
  group_by(mesocosm, date) %>%
  group_modify(~ calc_metabolism_day(
    .x, K600 = K600_median, depth = depth
  )) %>%
  ungroup()

# Method C: dynamic daily nighttime-regression estimate
results_nightreg_dynamic <- df %>%
  filter(!is.na(do_corrected), !is.na(temp_final)) %>%
  mutate(date = as.Date(timestamp)) %>%
  left_join(
    nightreg_clean %>% select(mesocosm, date, K600_nightreg),
    by = c("mesocosm", "date")
  ) %>%
  filter(!is.na(K600_nightreg)) %>%
  group_by(mesocosm, date, K600_nightreg) %>%
  group_modify(~ calc_metabolism_day(
    .x, K600 = unique(.y$K600_nightreg), depth = depth
  )) %>%
  ungroup()
```

## 5.3 Compare physiological signs and retained days

```{r compare_K600_methods}
summarise_K600_method <- function(data, method, K600_value) {
  data %>%
    summarise(
      `ER positive (%)` = round(mean(ER > 0, na.rm = TRUE) * 100, 1),
      `GPP negative (%)` = round(mean(GPP < 0, na.rm = TRUE) * 100, 1),
      `Retained mesocosm-days` = sum(!is.na(GPP) & !is.na(ER))
    ) %>%
    mutate(Method = method, K600 = K600_value, .before = 1)
}

K600_method_comparison <- bind_rows(
  summarise_K600_method(
    results_6.72,
    "Fixed literature K600",
    "6.72"
  ),
  summarise_K600_method(
    results_nightreg_median,
    "Fixed nighttime-regression median",
    sprintf("%.2f", K600_median)
  ),
  summarise_K600_method(
    results_nightreg_dynamic,
    "Dynamic daily nighttime regression",
    "Daily dynamic"
  )
)

K600_method_comparison %>%
  kable(caption = "Comparison of GPP and ER results among the three K600 methods")
```

# 5.4 Select K600 = 6.72 for the main analysis

The fixed literature value of K600 = 6.72 was selected as the primary method
because it provided the best balance between physiologically expected GPP and
ER signs and complete temporal coverage. All analyses below use this method
only.

```{r select_primary_K600}
results <- results_6.72

cat("Selected K600: 6.72 (fixed literature value)\n")
cat("Mesocosms:", length(unique(results$mesocosm)), "\n")
cat("Days:", length(unique(results$date)), "\n")
```

# 6. Visualisation

```{r gpp_er_plot_helpers}
# Prepare long-format data for GPP/ER plots
plot_data <- results %>%
  left_join(treatment_info, by = "mesocosm") %>%
  mutate(
    treatment  = factor(treatment, levels = TREATMENT_LEVELS),
    water_type = factor(water_type, levels = c("Ground", "River"))
  ) %>%
  pivot_longer(cols = c(GPP, ER), names_to = "metric", values_to = "value")

early_period <- seq(as.Date("2025-05-15"), as.Date("2025-05-22"), by = "day")

# Time series
plot_timeseries <- function(data_long) {
  ggplot(data_long, aes(x = date, y = value, color = metric)) +
    geom_line() +
    geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
    facet_wrap(~ mesocosm, ncol = 4) +
    scale_color_manual(values = METRIC_COLOURS) +
    labs(
      title = "Daily GPP and ER per Mesocosm (K600 = 6.72)",
      x     = "Date",
      y     = expression(g~O[2]~m^{-2}~d^{-1}),
      color = ""
    ) +
    theme_xinlin() +
    theme(axis.text.x = element_text(size = 7),
          strip.text  = element_text(size = 8))
}

# Boxplot by treatment (sqrt x-axis)
plot_boxplot <- function(data_long) {
  data_long %>%
    filter(!is.na(value)) %>%
    mutate(treatment = as.numeric(as.character(treatment))) %>%
    ggplot(aes(x = treatment, y = value, fill = metric,
               group = interaction(treatment, metric))) +
    geom_boxplot(outlier.shape = 21, outlier.size = 1.5) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
    facet_wrap(~ water_type, ncol = 1) +
    scale_x_continuous(trans = scales::transform_sqrt(), breaks = TREATMENT_LEVELS) +
    scale_fill_manual(values = METRIC_COLOURS) +
    labs(
      title = "GPP and ER by Basalt Treatment (K600 = 6.72)",
      x    = expression("ERW application (t ha"^{-1}*")"),
      y    = expression(g~O[2]~m^{-2}~d^{-1}),
      fill = ""
    ) +
    theme_xinlin()
}

```


```{r daily_nep_plot_helper}
# Daily NEP
plot_daily_NEP <- function(results) {
  results %>%
    left_join(treatment_info, by = "mesocosm") %>%
    mutate(
      NEP = GPP + ER,
      treatment = factor(treatment, levels = TREATMENT_LEVELS)
    ) %>%
    group_by(date, treatment, water_type) %>%
    summarise(
      mean_NEP = mean(NEP, na.rm = TRUE),
      se_NEP = sd(NEP, na.rm = TRUE) / sqrt(sum(!is.na(NEP))),
      .groups = "drop"
    ) %>%
    ggplot(aes(x = date, y = mean_NEP, color = treatment, fill = treatment)) +
    geom_ribbon(aes(ymin = mean_NEP - se_NEP,
                    ymax = mean_NEP + se_NEP),
                alpha = 0.08, color = NA) +
    geom_line(linewidth = 1) +
    geom_point(size = 0.8) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
    geom_vline(xintercept = as.Date("2025-05-22"),
               linetype = "dashed", color = "grey50") +
    facet_wrap(~ water_type, ncol = 2,
               labeller = as_labeller(c(Ground = "(a) Ground", River = "(b) River"))) +
    scale_color_viridis_d(option = "plasma", name = "ERW application (t/ha)") +
    scale_fill_viridis_d(option  = "plasma", name = "ERW application (t/ha)") +
    coord_cartesian(ylim = c(-1.3, NA)) +
    labs(
      title = "Daily NEP by Basalt Treatment (K600 = 6.72)",
      x = "Date",
      y = expression(Daily~NEP~(g~O[2]~m^{-2}~d^{-1}))
    ) +
    theme_xinlin() +
    theme(
      legend.position = "right",
      legend.direction = "vertical",
      legend.key.height = unit(0.5, "cm"),
      legend.spacing.y = unit(0.15, "cm"),
      strip.text = element_text(hjust = 0, size = 14),
      axis.text.x = element_text(size = 13, angle = 0, hjust = 0.5),
      axis.text.y = element_text(size = 13),
      axis.title.x = element_text(size = 14),
      axis.title.y = element_text(size = 14),
      legend.text = element_text(size = 14),
      legend.title = element_text(size = 14)
    )
}

```


# 6.1 Figures

## Time series

```{r fig_timeseries, fig.width=14, fig.height=10}
print(plot_timeseries(plot_data))

```

## Boxplot by treatment

```{r fig_boxplot, fig.width=8, fig.height=10}
print(plot_boxplot(plot_data))
```

## Daily NEP

```{r fig_daily_nep, fig.width=14, fig.height=6}
print(plot_daily_NEP(results))
```

## Maximum daily NEP by water type

The following summary reports the maximum treatment-level mean daily NEP for
each water type, together with the date and ERW treatment at which it occurred.
These values are calculated using the same grouping as the daily NEP figure.

```{r max_daily_nep_by_water}
daily_nep_summary <- results %>%
  left_join(treatment_info, by = "mesocosm") %>%
  mutate(NEP = GPP + ER) %>%
  group_by(water_type, treatment, date) %>%
  summarise(
    mean_NEP = mean(NEP, na.rm = TRUE),
    n = sum(!is.na(NEP)),
    .groups = "drop"
  )

max_daily_nep_by_water <- daily_nep_summary %>%
  filter(is.finite(mean_NEP)) %>%
  group_by(water_type) %>%
  slice_max(order_by = mean_NEP, n = 1, with_ties = TRUE) %>%
  ungroup() %>%
  mutate(
    water_type = dplyr::recode(
      as.character(water_type),
      Ground = "Groundwater",
      River = "River water"
    )
  ) %>%
  arrange(water_type, date, treatment)

max_daily_nep_by_water %>%
  select(water_type, mean_NEP, date, treatment, n) %>%
  kable(
    digits = 3,
    col.names = c(
      "Water type",
      "Maximum mean daily NEP (g O2 m-2 d-1)",
      "Date",
      "ERW treatment (t/ha)",
      "n"
    ),
    caption = "Maximum mean daily NEP for each water type"
  )

max_daily_nep_by_water %>%
  mutate(
    result = sprintf(
      "%s: the maximum mean daily NEP was %.3f g O2 m-2 d-1 on %s at the %s t/ha ERW treatment (n = %d).",
      water_type,
      mean_NEP,
      format(date, "%d %B %Y"),
      treatment,
      n
    )
  ) %>%
  pull(result) %>%
  cat(sep = "\n")
```

# 7. Linear mixed-effects models

Final model: `Response ~ treatment + water_type + (1|mesocosm) + (1|date)`
0 t/ha control included, no interaction term (selected by AIC).

## Prepare data and fit models

```{r lmm_fit}
prepare_lmm_data <- function(results) {
  base <- results %>%
    left_join(treatment_info, by = "mesocosm") %>%
    mutate(
      treatment  = as.numeric(as.character(treatment)),
      date       = as.factor(as.Date(date)),
      water_type = as.factor(water_type),
      mesocosm   = as.factor(mesocosm)
    )
  nep_data    <- base %>% filter(!is.na(GPP), !is.na(ER)) %>% mutate(NEP = GPP + ER)
  gpp_er_data <- base %>% filter(!is.na(GPP), !is.na(ER))
  list(
    nep_data               = nep_data,
    gpp_er_data            = gpp_er_data,
    nep_data_no_control    = nep_data    %>% filter(treatment != 0),
    gpp_er_data_no_control = gpp_er_data %>% filter(treatment != 0)
  )
}

d <- prepare_lmm_data(results)

f  <- function(resp, dat)
  lmer(as.formula(paste0(resp, " ~ treatment + water_type + (1|mesocosm) + (1|date)")),
       data = dat)
fi <- function(resp, dat)
  lmer(as.formula(paste0(resp, " ~ treatment + water_type + treatment:water_type + (1|mesocosm) + (1|date)")),
       data = dat)

# All four model variants for AIC comparison
M_NEP          <- fi("NEP", d$nep_data)
M_GPP          <- fi("GPP", d$gpp_er_data)
M_ER           <- fi("ER",  d$gpp_er_data)

M_NEP_final    <- f("NEP",  d$nep_data)
M_GPP_final    <- f("GPP",  d$gpp_er_data)
M_ER_final     <- f("ER",   d$gpp_er_data)

M_NEP_nc       <- fi("NEP", d$nep_data_no_control)
M_GPP_nc       <- fi("GPP", d$gpp_er_data_no_control)
M_ER_nc        <- fi("ER",  d$gpp_er_data_no_control)

M_NEP_nc_noint <- f("NEP",  d$nep_data_no_control)
M_GPP_nc_noint <- f("GPP",  d$gpp_er_data_no_control)
M_ER_nc_noint  <- f("ER",   d$gpp_er_data_no_control)
```

## AIC comparison

```{r lmm_aic}
aic_table <- tibble(
  Response = rep(c("NEP", "GPP", "ER"), each = 4),
  Model    = rep(c("Full (interaction, all data)",
                   "No interaction, all data — FINAL",
                   "No control, with interaction",
                   "No control, no interaction"), times = 3),
  AIC = round(c(
    AIC(M_NEP), AIC(M_NEP_final), AIC(M_NEP_nc), AIC(M_NEP_nc_noint),
    AIC(M_GPP), AIC(M_GPP_final), AIC(M_GPP_nc), AIC(M_GPP_nc_noint),
    AIC(M_ER),  AIC(M_ER_final),  AIC(M_ER_nc),  AIC(M_ER_nc_noint)
  ), 2)
)

aic_table %>% kable(caption = "AIC by model variant (compare within same Response only)")
```

## Fixed-effect coefficients (final model)

```{r lmm_coefficients}
coef_table <- bind_rows(
  tidy(M_NEP_final, effects = "fixed") %>% mutate(Response = "NEP"),
  tidy(M_GPP_final, effects = "fixed") %>% mutate(Response = "GPP"),
  tidy(M_ER_final,  effects = "fixed") %>% mutate(Response = "ER")
) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    estimate  = round(estimate,  4),
    std.error = round(std.error, 4),
    p.value   = round(p.value,   4),
    sig = case_when(
      p.value < 0.001 ~ "***", p.value < 0.01 ~ "**",
      p.value < 0.05  ~ "*",   p.value < 0.1  ~ ".",
      TRUE            ~ "ns"
    )
  ) %>%
  select(Response, term, estimate, std.error, p.value, sig)

coef_table %>%
  kable(caption = "Fixed-effect coefficients — final model (no interaction, 0 t/ha included)")
```

## Type II ANOVA

```{r lmm_anova}
anova_table <- bind_rows(
  anova(M_NEP_final, type = 2, ddf = "Satterthwaite") %>%
    as.data.frame() %>% rownames_to_column("term") %>% mutate(Response = "NEP"),
  anova(M_GPP_final, type = 2, ddf = "Satterthwaite") %>%
    as.data.frame() %>% rownames_to_column("term") %>% mutate(Response = "GPP"),
  anova(M_ER_final,  type = 2, ddf = "Satterthwaite") %>%
    as.data.frame() %>% rownames_to_column("term") %>% mutate(Response = "ER")
) %>%
  select(Response, term, `F value`, `Pr(>F)`) %>%
  mutate(
    `F value` = round(`F value`, 3),
    `Pr(>F)`  = round(`Pr(>F)`,  4),
    sig = case_when(
      `Pr(>F)` < 0.001 ~ "***", `Pr(>F)` < 0.01 ~ "**",
      `Pr(>F)` < 0.05  ~ "*",   `Pr(>F)` < 0.1  ~ ".",
      TRUE             ~ "ns"
    )
  )

anova_table %>%
  kable(caption = "Type II ANOVA — final model")
```

## R-squared

```{r lmm_rsquared}
rsq_table <- bind_rows(
  bind_cols(Response = "NEP", as.data.frame(r.squaredGLMM(M_NEP_final))),
  bind_cols(Response = "GPP", as.data.frame(r.squaredGLMM(M_GPP_final))),
  bind_cols(Response = "ER",  as.data.frame(r.squaredGLMM(M_ER_final)))
) %>%
  mutate(across(where(is.numeric), ~ round(.x, 3)))

rsq_table %>%
  kable(caption = "Marginal (R2m) and conditional (R2c) R-squared — final model")
```

## 7.1 Daily NEP model with date as a fixed effect

This additional model treats sampling date as a categorical fixed effect:
`NEP ~ treatment + water_type + date + (1|mesocosm)`.

```{r daily_nep_date_fixed_model}
M_NEP_date_fixed <- lmer(
  NEP ~ treatment + water_type + date + (1 | mesocosm),
  data = d$nep_data
)

# Complete model summary
summary(M_NEP_date_fixed)

# Fixed-effect coefficient results
nep_date_fixed_coefficients <- tidy(
  M_NEP_date_fixed,
  effects = "fixed",
  conf.int = TRUE
) %>%
  mutate(across(where(is.numeric), ~ round(.x, 4)))

nep_date_fixed_coefficients %>%
  kable(caption = "Daily NEP model coefficients — date fitted as a fixed effect")

# Overall tests for treatment, water type and date
nep_date_fixed_anova <- anova(
  M_NEP_date_fixed,
  type = 2,
  ddf = "Satterthwaite"
) %>%
  as.data.frame() %>%
  rownames_to_column("term") %>%
  mutate(across(where(is.numeric), ~ round(.x, 4)))

nep_date_fixed_anova %>%
  kable(caption = "Type II ANOVA — daily NEP model with date as a fixed effect")
```

## 7.2 Model diagnostics (DHARMa)

```{r lmm_diagnostics, fig.width=8, fig.height=6}
for (resp in c("NEP", "GPP", "ER")) {
  model <- get(paste0("M_", resp, "_final"))
  cat("\n\n**", resp, "diagnostics**\n\n")
  sim <- simulateResiduals(model)
  plot(sim, main = paste("Final model —", resp))
}
```

## 7.3 Prediction plots

```{r lmm_predictions, fig.width=16, fig.height=6}
make_pred_panel <- function(pred, raw_data, y_var, y_label, title) {
  pred %>%
    as.data.frame() %>%
    mutate(panel_label = title) %>%
    ggplot(aes(x = x, y = predicted, colour = group, fill = group)) +
    geom_point(
      data = raw_data %>% mutate(panel_label = title),
      aes(x = treatment, y = .data[[y_var]], colour = water_type),
      inherit.aes = FALSE,
      alpha = 0.15,
      size = 1
    ) +
    geom_ribbon(
      aes(ymin = conf.low, ymax = conf.high),
      alpha = 0.2,
      colour = NA
    ) +
    geom_line(linewidth = 1) +
    geom_hline(
      yintercept = 0,
      linetype = "dashed",
      colour = "grey50"
    ) +
    facet_wrap(~ panel_label) +
    scale_x_continuous(
      trans = scales::transform_sqrt(),
      breaks = TREATMENT_LEVELS
    ) +
    scale_colour_manual(
      values = WATER_COLOURS,
      labels = WATER_LABELS
    ) +
    scale_fill_manual(
      values = WATER_COLOURS,
      labels = WATER_LABELS
    ) +
    labs(
      x = expression("ERW application (t ha"^{-1}*")"),
      y = y_label,
      colour = "Water type",
      fill = "Water type"
    ) +
    theme_classic() +
    theme(
      legend.position = "right",
      strip.background = element_blank(),
      strip.text = element_text(
        face = "bold",
        hjust = 0,
        size = 14
      ),
      axis.text.x = element_text(
        hjust = 1,
        size = 13
      ),
      axis.text.y = element_text(
        size = 13
      ),
      axis.title.x = element_text(
        size = 14
      ),
      axis.title.y = element_text(
        size = 14
      ),
      legend.text = element_text(
        size = 14
      ),
      legend.title = element_text(
        size = 14
      )
    ) +
    geom_jitter(
      data = raw_data %>% mutate(panel_label = title),
      aes(
        x = treatment,
        y = .data[[y_var]],
        colour = water_type
      ),
      inherit.aes = FALSE,
      alpha = 0.15,
      size = 1,
      width = 0.8,
      height = 0
    )
}

pred_NEP <- ggpredict(
  M_NEP_final,
  terms = c("treatment [0,5,10,20,50,100]", "water_type")
)

pred_GPP <- ggpredict(
  M_GPP_final,
  terms = c("treatment [0,5,10,20,50,100]", "water_type")
)

pred_ER <- ggpredict(
  M_ER_final,
  terms = c("treatment [0,5,10,20,50,100]", "water_type")
)

p_NEP <- make_pred_panel(
  pred_NEP,
  d$nep_data,
  "NEP",
  expression(NEP~(g~O[2]~m^{-2}~d^{-1})),
  "(a)"
)

p_GPP <- make_pred_panel(
  pred_GPP,
  d$gpp_er_data,
  "GPP",
  expression(GPP~(g~O[2]~m^{-2}~d^{-1})),
  "(b)"
)

p_ER <- make_pred_panel(
  pred_ER,
  d$gpp_er_data,
  "ER",
  expression(ER~(g~O[2]~m^{-2}~d^{-1})),
  "(c)"
)

p_combined <- p_NEP + p_GPP + p_ER +
  plot_layout(ncol = 3, guides = "collect") &
  theme(legend.position = "right")

p_combined
```


# 8. Save outputs

```{r save_outputs}
ggsave(file.path(params$output_dir, "GPP_ER_timeseries.png"),
       plot_timeseries(plot_data), width = 14, height = 10, dpi = 300)

ggsave(file.path(params$output_dir, "GPP_ER_boxplot.png"),
       plot_boxplot(plot_data), width = 8, height = 10, dpi = 300)

ggsave(file.path(params$output_dir, "daily_NEP.png"),
       plot_daily_NEP(results), width = 13, height = 6, dpi = 300)


ggsave(file.path(params$output_dir, "model_pred_combined.png"),
       p_combined, width = 17, height = 6, dpi = 300)

write_csv(results,      file.path(params$output_dir, "metabolism_results_K600_6.72.csv"))
write_csv(K600_estimation_summary,
          file.path(params$output_dir, "K600_nighttime_regression_summary.csv"))
write_csv(K600_method_comparison,
          file.path(params$output_dir, "K600_method_screening_comparison.csv"))
write_csv(nightreg_results,
          file.path(params$output_dir, "K600_nighttime_regression_all.csv"))
write_csv(nightreg_clean,
          file.path(params$output_dir, "K600_nighttime_regression_retained.csv"))
write_csv(aic_table,    file.path(params$output_dir, "lmm_AIC_comparison.csv"))
write_csv(coef_table,   file.path(params$output_dir, "lmm_coefficients_final_model.csv"))
write_csv(anova_table,  file.path(params$output_dir, "lmm_type2_anova_final_model.csv"))
write_csv(rsq_table,    file.path(params$output_dir, "lmm_Rsquared_final_model.csv"))
write_csv(nep_date_fixed_coefficients,
          file.path(params$output_dir, "daily_NEP_date_fixed_coefficients.csv"))
write_csv(nep_date_fixed_anova,
          file.path(params$output_dir, "daily_NEP_date_fixed_type2_anova.csv"))
```

# 9. Reproducibility

```{r session_info}
sessionInfo()
```
